---
title: "GCD by Herminte Normal Form"
output: github_document
date: "2024-11-17"
---


Extended GCD as Herminte normal form version of [the remainder problem](https://github.com/WinVector/Examples/blob/main/puzzles/DividingCoconuts/RemainderProblem.ipynb).

```{r}
# extended GCD by Hermite normal form
library(numbers)

a = matrix(c(
  723217 - 480608, 508811 - 480608),
  nrow = 1, ncol = 2, byrow = TRUE)

a
```

```{r}
hnf <- hermiteNF(a)

hnf
```

```{r}
stopifnot(all(a %*% hnf$U == hnf$H))

hnf$H  # GCD
```

```{r}
hnf$U[ , 1]  # multipliers
```

```{r}
hnf$U[1 , 1] * a[1, 1] + hnf$U[2 , 1] * a[1, 2]
```

Links:

  * [W. A. Blankinship, "A New Version of the Euclidean Algorithm", The American Mathematical Monthly, Vol. 70, No. 7 (Aug. - Sep., 1963), Taylor & Francis, Ltd. for Mathematical Association of America, pp. 742-745](https://www.jstor.org/stable/2312260).
  * George Havas, Bohdan S. Majewski, and Keith R. Matthews, "Extended GCD and Hermite Normal Form Algorithms via Lattice Basis Reduction", Experimental Mathematics 7:2, K Peters, Ltd. 058-6458/1998, pp. 1-25.
  * [S. A. Rankin, "The Euclidean Algorithm and the Linear Diophantine Equation ax + by = gcd(a, b)", The American Mathematical Monthly, Vol. 120, No. 6 (June–July 2013), pp. 562-564](https://www.jstor.org/stable/10.4169/amer.math.monthly.120.06.562).

