---
title: "Why replyr"
output: html_document
---



Why try `replyr`?  Because it lets you take a number of commmon working patterns and apply them to remote data (such as databases or `Spark`).  Below are some examples.

```{r setup}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
library("tidyr")
packageVersion("tidyr")
library("replyr")
packageVersion("replyr")
suppressPackageStartupMessages("spaklyr")
packageVersion("sparklyr")

sc <- sparklyr::spark_connect(version='2.0.2', 
                              master = "local")
```

## `dplyr::do`

Taking a few rows from each group of a grouped data set. 
Note: since we are not enforcing order by an arrange we 
can't expect the results to always match on database
or `Spark` data sources.

### `dplyr::do` on local data

From `help('do', package='dplyr')`:

```{r dplyrdolocal}
by_cyl <- group_by(mtcars, cyl)
do(by_cyl, head(., 2))
```

### `dplyr::do` on `Spark`

```{r dplyrdolocalspark}
mtcars_spark <- copy_to(sc, mtcars)
by_cyl <- group_by(mtcars_spark, cyl)
do(by_cyl, head(., 2))
```

Notice we did not get back usable results.

### `replyr` split/apply

```{r replyrdo}
mtcars_spark %>%
  replyr_split('cyl', 
               partitionMethod = 'extract') %>%
  lapply(function(di) head(di, 2)) %>%
  replyr_bind_rows()
```

### `replyr` gapply

```{r replyrgapply}
mtcars_spark %>%
  gapply('cyl',
         partitionMethod = 'extract',
         function(di) head(di, 2))
```

## Handle management



```{r cleanup}
sparklyr::spark_disconnect(sc)
rm(list=ls())
gc()
```
