---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# Why `replyr`

`replyr` stands for **RE**mote **PLY**ing of big data for **R**.

Why should [R](https://www.r-project.org) users try [`replyr`](https://CRAN.R-project.org/package=replyr)?  Because it lets you take a number of common working patterns and apply them to remote data (such as databases or [`Spark`](https://spark.apache.org)). 

`replyr` allows users to work with `Spark` or database data similar to how they work with local `data.frame`s. Some key capability gaps remedied by `replyr` include:

  * Summarizing data: `replyr_summary()`.
  * Combining tables: `replyr_union_all()`.
  * Binding tables by row: `replyr_bind_rows()`.
  * Using the split/apply/combine pattern (`dplyr::do()`): `replyr_split()`, `replyr::gapply()`.
  * Pivot/anti-pivot (`gather`/`spread`): `replyr_moveValuesToRows()`/ `replyr_moveValuesToColumns()`.
  * [Handle tracking](http://www.win-vector.com/blog/2017/06/managing-intermediate-results-when-using-rsparklyr/).
  * A [join controller](http://www.win-vector.com/blog/2017/06/use-a-join-controller-to-document-your-work/).

You may have already learned to decompose your local data processing into steps including the above,
so retaining such capabilities makes working with `Spark` and [`sparklyr`](http://spark.rstudio.com) *much* easier.
Some of the above capabilities will [likely come to the `tidyverse`](https://github.com/tidyverse/dplyr/issues/2923), but the above implementations are build purely on
top of `dplyr` and are the ones [already being vetted and debugged](https://github.com/rstudio/sparklyr/issues/783) at production scale (I think these will be ironed out and reliable sooner).

[`replyr`](https://github.com/WinVector/replyr) is the product of collecting experience applying `R` at production scale with many clients, collecting issues and filling-in gaps.

Below are some examples.

***

# Examples

***

Things are changing fast right now, so let's use the development versions of the packages for our examples.

```{r setup}
base::date()
# devtools::install_github('rstudio/sparklyr')
# devtools::install_github('tidyverse/dplyr')
# devtools::install_github('tidyverse/dbplyr')
# devtools::install_github('WinVector/replyr')
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
packageVersion("dbplyr")
library("tidyr")
packageVersion("tidyr")
library("replyr")
packageVersion("replyr")
suppressPackageStartupMessages(library("sparklyr"))
packageVersion("sparklyr")

# more memory as suggested in https://github.com/rstudio/sparklyr/issues/783
config <- spark_config()
config[["sparklyr.shell.driver-memory"]] <- "8G"
sc <- sparklyr::spark_connect(version='2.1.0', 
                              hadoop_version = '2.7',
                              master = "local",
                              config = config)
```

## `summary`

Standard `summary()` and `glance()`, all fail on `Spark`.

```{r sparksummary, error=TRUE}
mtcars_spark <- copy_to(sc, mtcars)

# gives summary of handle, not data
summary(mtcars_spark)
```

```{r glance, error=TRUE}
packageVersion("broom")
broom::glance(mtcars_spark)
```

`replyr_summary` works.

```{r rsum}
replyr_summary(mtcars_spark) %>%
  select(-lexmin, -lexmax, -nunique, -index)
```

***

## `gather`/`spread`

`tidyr` pretty much only works on local data.

```{r gatherspread, error=TRUE}
mtcars2 <- mtcars %>%
  mutate(car = row.names(mtcars)) %>%
  copy_to(sc, ., 'mtcars2')

# errors out
mtcars2 %>% 
  tidyr::gather('fact', 'value')
```



```{r movevaluestorows, eval=TRUE}
mtcars2 %>%
  replyr_moveValuesToRows(nameForNewKeyColumn= 'fact', 
                          nameForNewValueColumn= 'value', 
                          columnsToTakeFrom= colnames(mtcars),
                          nameForNewClassColumn= 'class') %>%
  arrange(car, fact)
```



## Binding rows

`dplyr` `bind_rows`, `union`, and `union_all` are all currently unsuitable for use on `Spark`.
`replyr::replyr_union_all()` and `replyr::replyr_bind_rows()` supply working alternatives.

### `bind_rows()`


```{r bindrows, error=TRUE}
db1 <- copy_to(sc, 
               data.frame(x=1:2, y=c('a','b'), 
                          stringsAsFactors=FALSE),
               name='db1')
db2 <- copy_to(sc, 
               data.frame(y=c('c','d'), x=3:4, 
                          stringsAsFactors=FALSE),
               name='db2')

# Errors out as it tries to operate on the handles instead of the data.
bind_rows(list(db1, db2))
```

### `union_all`

```{r uniona, error=TRUE}
# ignores column names and converts all data to char
union_all(db1, db2)
```

### `union`

```{r union, error=TRUE}
# ignores column names and converts all data to char
# also will probably lose duplicate rows
union(db1, db2)
```

### `replyr_bind_rows`

`replyr::replyr_bind_rows` can bind multiple `data.frame`s together.

```{r replyr_bind_rows, error=TRUE}
replyr_bind_rows(list(db1, db2))
```

## `dplyr::do`

Our example is just taking a few rows from each group of a grouped data set. 
Note: since we are not enforcing order by an arrange we 
can't expect the results to always match on database
or `Spark` data sources.

### `dplyr::do` on local data

From `help('do', package='dplyr')`:

```{r dplyrdolocal}
by_cyl <- group_by(mtcars, cyl)
do(by_cyl, head(., 2))
```

***

### `dplyr::do` on `Spark`

```{r dplyrdolocalspark}
by_cyl <- group_by(mtcars_spark, cyl)
do(by_cyl, head(., 2))
```

Notice we did not get back usable results.

### `replyr` split/apply

```{r replyrdo}
mtcars_spark %>%
  replyr_split('cyl', 
               partitionMethod = 'extract') %>%
  lapply(function(di) head(di, 2)) %>%
  replyr_bind_rows()
```

### `replyr` `gapply`

```{r replyrgapply}
mtcars_spark %>%
  gapply('cyl',
         partitionMethod = 'extract',
         function(di) head(di, 2))
```


## `replyr::replyr_apply_f_mapped`

[`wrapr::let`](http://www.win-vector.com/blog/2017/06/please-consider-using-wraprlet-for-replacement-tasks/) was only the secondary proposal in the original [2016 "Parametric variable names" article](http://www.win-vector.com/blog/2016/12/parametric-variable-names-and-dplyr/).  What we really wanted was a stack of view so the data pretended to have names that matched the code (i.e., re-mapping the data, not the code).  

With a bit of thought we can achieve this if we associate the data re-mapping with a function environment instead of with the data.  So a re-mapping is active as long as a given controlling function is in control.  In our case that function is `replyr::replyr_apply_f_mapped()` and works as follows:

Suppose the operation we wish to use is a rank-reducing function that has been supplied as function from somewhere else that we do not have control of (such as a package).  The function could be simple such as the following, but we are going to assume we want to use it without alteration (including the without the small alteration of introducing `wrapr::let()`).

```{r rankfn}
# an external function with hard-coded column names
DecreaseRankColumnByOne <- function(d) {
  d$RankColumn <- d$RankColumn - 1
  d
}
```

To apply this function to `d` (which doesn't have the expected column names!) we use  `replyr::replyr_apply_f_mapped()` to create a new parameterized adapter as follows:

```{r replyrmapf}
# our data
d <- data.frame(Sepal_Length = c(5.8,5.7),
                Sepal_Width = c(4.0,4.4),
                Species = 'setosa',
                rank = c(1,2))

# a wrapper to introduce parameters
DecreaseRankColumnByOneNamed <- function(d, ColName) {
  replyr::replyr_apply_f_mapped(d, 
                                f = DecreaseRankColumnByOne, 
                                nmap = c(RankColumn = ColName),
                                restrictMapIn = FALSE, 
                                restrictMapOut = FALSE)
}

# use
dF <- DecreaseRankColumnByOneNamed(d, 'rank')
print(dF)
```

`replyr::replyr_apply_f_mapped()` renames the columns to the names expected by `DecreaseRankColumnByOne` (the mapping specified in `nmap`), applies `DecreaseRankColumnByOne`, and then inverts the mapping before returning the value.

***

## Handle management

Many [`Sparklyr`](https://CRAN.R-project.org/package=sparklyr) tasks involve creation of intermediate or temporary tables.  This can be through `dplyr::copy_to()` and through `dplyr::compute()`. These handles can represent a reference leak and eat up resources.   

To help control handle lifetime the [`replyr`](https://CRAN.R-project.org/package=replyr) supplies handle trackers: record-retaining temporary name generators (and uses the same internally).

The actual function is pretty simple:

```{r printtr}
print(replyr::makeTempNameGenerator)
```

For instance to join a few tables it can be a good idea to call `compute` after each join for some data sources (else the generated `SQL` can become large and unmanageable).  This sort of code looks like the following (now hanging with `sparklyr `0.5.6.9003` June 22, 2017):

```{r joinlist, eval=TRUE}
# create example data
names <- paste('table', 1:5, sep='_')
tables <- lapply(names, 
                 function(ni) {
                   di <- data.frame(key= 1:3)
                   di[[paste('val',ni,sep='_')]] <- runif(nrow(di))
                   copy_to(sc, di, ni)
                 })

# build our temp name generator
tmpNamGen <- replyr::makeTempNameGenerator('JOINTMP')

# left join the tables in sequence
joined <- tables[[1]]
for(i in seq(2,length(tables))) {
  ti <- tables[[i]]
  if(i<length(tables)) {
    joined <- compute(left_join(joined, ti, by='key'),
                    name= tmpNamGen())
  } else {
    # use non-temp name.
    joined <- compute(left_join(joined, ti, by='key'),
                    name= 'joinres')
  }
}

# clean up temps
temps <- tmpNamGen(dumpList = TRUE)
print(temps)
for(ti in temps) {
  db_drop_table(sc, ti)
}

# show result
print(joined)
```

Careful introduction and management of materialized intermediates can conserve resources (both time and space) and greatly improve outcomes.  We feel it is a good practice to set up an explicit temp name manager, pass it through all your `Sparklyr` transforms, and then clear temps in batches after the results no longer depend on the intermediates.

***

# Conclusion

If you are serious about `R` controlled data processing in `Spark` or databases you should seriously consider using `replyr` in addition to [`dplyr`](https://CRAN.R-project.org/package=dplyr) and `sparklyr`.

Be aware of the functionality we demonstrated depends on using the development version of `replyr`.  Though we will, of course, advance the CRAN version as soon as practical. 

***

```{r cleanup}
sparklyr::spark_disconnect(sc)
rm(list=ls())
gc()
```
