---
title: "The Perils of Prevalence Adjustment"
output: github_document
---

## Introduction

I would like to show how the *point-wise shift model homotopy as <code>P</code>* is not unbiased and not the same as the *unbiased shift model homotopy defined as <code>U</code>*. We define a [probability model homotopy](https://win-vector.com/2020/10/10/upcoming-series-probability-model-homotopy/) as a set of related models addressed by an index variable [here](https://win-vector.com/2020/10/10/upcoming-series-probability-model-homotopy/). In the context of this note it is enough to think of a model homotopy as family of models related correction scheme we intend to apply when the model is applied to new data.

In our introductory note we mentioned a few common corrections.

  * The <code>P</code>-correction, which is essentially a point-wise application of Bayes' Law.
  * The <code>U</code>-correction, which is designed to get prevalence right.
  
Our point is, these are not the same correction. Thus the <code>P</code>-correction is not always unbiased.

### Example Setup

Let's demonstrate this in [<code>R</code>](https://www.r-project.org).

First we attach our packages.

```{r}
library(wrapr)
library(numbers)
```

Now let's build our simulated model performance data.

```{r}
bal_size <- 8
n_imb_rep <- 80
d <- data.frame(
  prediction = c(
    rep((bal_size - 1)/bal_size, bal_size),
    rep(c(1/2, 1/2, 1/2, 1/4, 1/4, 1/4, 1/4, 1/4, 1/4), n_imb_rep),
    rep(1/bal_size, bal_size)),
  truth = c(
    rep(TRUE, (bal_size - 1)), FALSE,
    rep(c(TRUE, TRUE, FALSE, TRUE, rep(FALSE, 5)), n_imb_rep),
    TRUE, rep(FALSE, (bal_size - 1))
  ))
d$orig_row_id <- seq_len(nrow(d))

knitr::kable(head(d))
```

The column `prediction` is behaving like an unbiased probability model predicting `truth`. In particular `prediction` and `truth` have the same expected value on this data set.

```{r}
colMeans(subset(subset(d, select= -orig_row_id)))
```

```{r}
prevalence <- mean(d$truth)

prevalence
```

```{r}
epsilon <- 1.0e-9
stopifnot(abs(prevalence  -  mean(d$prediction)) < epsilon)
prevalence  -  mean(d$prediction)
```

## Bayes' Correction (<code>P</code>)

Now the idea is the <code>P</code> correction is as follows. By Bayes' Law we have:

```
  P[outcome==TRUE | evidence] = 
     P[outcome==TRUE] P[evidence | outcome==TRUE] / P[evidence]
     
  (1 - P[outcome==TRUE | evidence]) = 
     (1 - P[outcome==TRUE]) P[evidence | outcome==FALSE] / P[evidence]
```

So in terms of odds-ratios we have:

```
 P[outcome==TRUE | evidence] / (1 - P[outcome==TRUE | evidence]) = 
    (P[outcome==TRUE] / (1 - P[outcome==TRUE])) *
    (P[evidence | outcome==TRUE] / P[evidence | outcome==FALSE])
```

Taking logs of both-sides allows us to re-write this in terms of `logit`.


```{r}
logit <- function(x) {
  log( x / (1 - x) )
}
```

Which gives us

```
  logit(P[outcome==TRUE | evidence]) = 
     logit(P[outcome==TRUE]) + log(P[evidence | outcome==TRUE] / P[evidence | outcome==FALSE])
```

The idea is:

  * `log(P[evidence | outcome==TRUE] / P[evidence | outcome==FALSE])` is independent of the outcome prevalence of the data set it was estimated on.
  * `logit(P[outcome==TRUE])` is a function of a prevalence.
  
So it is tempting to take only the `log(P[evidence | outcome==TRUE] / P[evidence | outcome==FALSE])` term from our model, and training data and plug in an estimate of the prevalence of the population we intend to work with. For example we showed in ["A Gruesome Example of Bayes’ Law"](https://win-vector.com/2020/09/11/a-gruesome-example-of-bayes-law/) how to take an conditional evidence odds-ratio built on a population with a prevalence near 50%, and apply it to a population with low disease prevalence. Without this correction the predictions would be very far off!

The Bayes/<code>P</code> correction is developed as follows.

```{r}
sigmoid <- function(x) {
  1 / (1 + exp(-x))
}
```

Suppose we intend to apply this model, built on a population with `r format(prevalence, digits=2)` prevalence on a new population with 50% prevalence. Let's suppose our 50% prevalence is just a deterministic even re-sampling of our data (a very easy case!.

```{r}
# get how many times to replicate each row group
count_table <- aggregate(
  count ~ truth, 
  data = transform(d, count = 1), 
  FUN = length)

multiple <- Reduce(LCM, count_table$count)
multiple
```

```{r}
count_table$n_reps <- multiple / count_table$count

knitr::kable(count_table)
```

```{r}
# replicate each row group by the target number of times
rep_table <- count_table %.>%
 lapply(
  seq_len(nrow(.)),
  function(i) {
    data.frame(
      truth = .$truth[[i]],
      row_rep = seq_len(.$n_reps[[i]]))
  }) %.>%
  do.call(rbind, .)

d_2 <- merge(d, rep_table, by = 'truth') %.>%
  .[order(.$orig_row_id, .$row_rep), ]
rownames(d_2) <- NULL

knitr::kable(head(d_2))

d_2$orig_row_id <- NULL
d_2$row_rep <- NULL
```


```{r}
prevalence_2 <- mean(d_2$truth)

prevalence_2
```

Now notice our original uncorrected model does not match the prevalence on this new population. It is off and behaving in a biased manner.

```{r}
stopifnot(abs(prevalence_2  - mean(d_2$prediction)) > 1e-2)
mean(d_2$prediction)
```

The Bayes/<code>P</code> correction is given by shifting the model in logit/link-space by the following factor.


```{r}
delta <- -logit(prevalence) + logit(prevalence_2)

delta
```


```{r}
d_2$p_adjusted_prediction <- sigmoid(
  logit(d_2$prediction) + delta)
```

And here is a summary of what we have. The `prediction` column is original model's predictions, the `truth` column is now the expected value of the truth indicator on the new re-sampled data set and the `p_adjusted_prediction` is our new Bayes adjusted prediction.

```{r}
aggregate(. ~ prediction, data = d_2, FUN = mean) %.>%
  knitr::kable(.)
```

And we have our problem, both the original and adjusted prediction remain biased.

```{r}
colMeans(d_2)
```

```{r}
stopifnot(abs(prevalence_2  - mean(d_2$p_adjusted_prediction)) > 1e-2)
mean(d_2$p_adjusted_prediction)
```


Neither of them equals the actual prevalence of 0.5, though the corrected value is in this case closer.

### Why the adjusted model is still biased

If the original model was unbiased (which it was), and the adjustment is correct (which it is), how can the result be biased (which it is)?

The answer is: the adjustment is unbiased for *perfect* models, which are not the kind we see in practice. A perfect model has absolutely no structure in the residuals. That is a perfect model has `P[outcome==TRUE | prediction = p] = p` for *all* <code>p</code>. This wasn't the case
for our example model. Here are the expected values of truth on the original un-weighted data set conditioned on the predicted probability.

```{r}
d %.>%
  subset(., select = -orig_row_id) %.>%
  aggregate(
    . ~ prediction, 
    data = ., 
    FUN = mean) %.>%
  knitr::kable(.)
```

Notice that prediction groups that were perfect in the original (`0.125`, and `0.875`) do match the truth expectation in the re-sampled data set. The idea is in this case the resampling is the exact same adjustment as the shift, so things that matched before adjustment match after adjustment.

For prediction groups that were not perfect in the original (`0.250` and `0.500`) the re-scaling of data moves the answer proportional the actual prevelance seen and the shifting of the prediction operates at the different rate predicted by the model. The sigmoid has different shapes in different regions, so the two estimates move at different rates. As we have seen, groups that compensated for each other in the original may fall out of sync and not compensate for each other after transform. The `0.250` prediction group was over-predicting in the original data frame, and the `0.500` groups was under-predicting in the original data frame. At the exact counts in the original data frame these two erros canceled, after transform the cancellation was maintained.

The point is: the idea that the Bayes' correction gets the expected value right is something we have to check. The obvious way to establish it would be to insist the mode obey *very* strong additional per-prediction range balance conditions. These conditions are typically not all met in production, unless one has taken extra care to [calibrate the model](https://en.wikipedia.org/wiki/Platt_scaling) (perhaps through isotonic regression, but even this will fail if the model has order-reversals; binning the predictions and using a nested model can fix the issue in general).


## The unbiased <code>U</code> correction.

Let's explore a correction whose only goal is to get the global average right. We hope it does more, but this is all it is really designed to do. What we do is apply a shift in link space similar to the last section, but we search for a shift that gets expected values right.

```{r}
f <- function(d) {
  mean(d_2$truth) - mean(sigmoid(logit(d_2$prediction) + d))
}

delta_2 <- uniroot(f, c(-2, 2), tol = .Machine$double.eps)

delta_2
```

This lands a new set of adjusted predictions.

```{r}
d_2$u_adjusted_prediction <- sigmoid(
  logit(d_2$prediction) + delta_2$root)

aggregate(. ~ prediction, data = d_2, FUN = mean) %.>%
  knitr::kable(.)
```

And these predictions are, as we hoped mean `0.5`.

```{r}
stopifnot(abs(prevalence_2  -  mean(d_2$u_adjusted_prediction)) < epsilon)
mean(d_2$u_adjusted_prediction)
```

## Platt Scaling

We can also try [Platt scaling](https://en.wikipedia.org/wiki/Platt_scaling) and a Platt-shift (no scaling, just a new intercept term).

As we expect from [knowledge of logistic regression](https://win-vector.com/2011/09/14/the-simpler-derivation-of-logistic-regression/), we see Platt scaling is unbiased in that it gets the global average right.

```{r}
Platt_scaler <- glm(
  truth ~ logit(prediction), 
  data = d_2, 
  family = binomial())

d_2$Platt_scaled_prediction <- predict(
  Platt_scaler,
  newdata = d_2,
  type = 'response')

aggregate(. ~ prediction, data = d_2, FUN = mean) %.>%
  knitr::kable(.)
```

```{r}
stopifnot(abs(prevalence_2  -  mean(d_2$Platt_scaled_prediction)) < epsilon)
mean(d_2$Platt_scaled_prediction)
```

Platt shifting work similarly, with one fewer degree of freedom.

```{r}
Platt_shifter <- glm(
  truth ~ 1, 
  offset = logit(prediction),
  data = d_2, 
  family = binomial())

d_2$Platt_shifted_prediction <- predict(
  Platt_shifter,
  newdata = d_2,
  type = 'response')

aggregate(. ~ prediction, data = d_2, FUN = mean) %.>%
  knitr::kable(.)
```


```{r}
stopifnot(abs(prevalence_2  -  mean(d_2$Platt_shifted_prediction)) < epsilon)
mean(d_2$Platt_shifted_prediction)
```

And we see Platt-shifting is exactly the <code>U</code> correction.

## Conclusion

Throughout this note we used "correction" instead of "model homotopy" to keep things conversational. There is a downside to this as when using the term "correction" it is awkward to discuss incorrect corrections, as one runs into the specious argument that wrong corrections are not corrections, so there is no such thing as wrong corrections. With "model homotopy" we separate what the homotopy is, from what was intended, and from how it is implemented. We can slow down and look at proposals using this notation.

The point of this note was: Bayesian corrections are not necessarily unbiased. This should not be too surprising as insisting on un-biasedness is a largely frequentest foundation. We have demonstrated, rhetoric chains of the form "this is the right correction", "a perfect correction would have this additional property", "therefore this correction would also have this property" can fall apart.

