---
title: "Some Examples Where re-Weighting Data Doesn't do Much"
author: "John Mount, Nina Zumel; https://www.win-vector.com"
date: "`r date()`"
output: github_document
---


## Introduction

Re-weighting data for unbalanced classification tasks is a popular data transform used to attempt to improve model sensitivity for important rare events. The method can help, but [we feel it is over-used as it also helps to work around the common mistake of using classification rules where numeric or probability models are more appropriate](https://win-vector.com/2020/08/07/dont-use-classification-rules-for-classification-problems/). 

[Nina Zumel had some interesting comments on the ideas](https://ninazumel.com/2015/02/27/balancing-classes-before-training-classifiers-addressing-a-folk-theorem/) which lead us to conclude: if re-balancing does anything better than moving your threshold, this is in fact evidence of a missed interaction.

It is our thesis that their is little benefit to re-balancing data and if there appears to be such a benefit it means you failded to use numeric scores (converted to a classification rule too early) or missed an interaction in your data (which can be fixed by a bit more feature engineering, the non-montone change suggests some interactions that can be introduced).

Let's work some simple examples to try and illustrate this point.

### Terminology

In this note (and our earlier one: ["Don’t Use Classification Rules for Classification Problems"](https://win-vector.com/2020/08/07/dont-use-classification-rules-for-classification-problems/)) we are going to need to lock down some terms at very specific uses (some standard, some a little less standard).

Our problem is: modeling an outcome or dependent variable that takes the value `TRUE`/`FALSE` (or `1`/`0`) in terms of numeric explanatory variables. A *numeric score model* is a model that returns a number or score for each instance, with higher scores being predictions that the dependent variable is more likely to be `TRUE`. A *probability model* is a score model where the number returned is in the range `[0, 1]` *and* we expect the probability that the explanatory is `TRUE` (over some aggregation) *is* estimated by the model score.

A *classification rule* is a model that returns a decision for each instance: a prediction of "POSITIVE" indicating a prediction that the dependent is `TRUE` and "NEGATIVE" indicating a prediction that the dependent variables is `FALSE`. Often a classification rule is formed by taking a scoring model and declaring the determination to be "POSITIVE" if the score is at least a value called the *classification threshold*.

In [`R`](https://www.r-project.org) the standard logistic regression `stats::glm()` *always* returns numeric scores (be they "link", "response", or even "terms"). So `R` users tend to use numeric scores.

In [`Python`](https://www.python.org) [`scikit-learn`](https://scikit-learn.org/stable/) the convention is `.predict()` returns categorical decisions, and `.predict_proba()` returns scores (usually probabilities). So `Python` users are led toward classification rules as `.predict()` is the more common method. However, the mere discipline of going to the trouble to use `.predict_proba()` disposes of this bias.

The type of *data re-weighting* we are discussing is "outcome proportional": all of the weights of one outcome class (usually the positive examples) are re-weighted by the same constant. Common examples of this are "positive up-sampling" (replicating positive examples to give them more emphasis), "negative down-sampling" (down-weighting or censoring negative examples to give them less emphasis). The issues of these procedures: up-sampling breaks example independence, up-sampling can cause data leaks if examples are not moved in groups during cross-validation and test/train split, down-sampling loses some statistical efficiency, and both procedures no longer match the actual training distribution prevalence.

Obviously arbitrary data re-weighting (changing each row's amount of representation in the data set arbitrarily) can change everything about the model, as this has enough power to select completely different sets of training data under different re-weightings. The "outcome proportional" re-weighting, is are more limited set of transformations.

[Our thesis is](https://win-vector.com/2020/08/07/dont-use-classification-rules-for-classification-problems/): a model that returns a continuous numeric score typically contains *more* useful information than any one classification rule. One should work with scores as much as possible, and delay the conversion to rules as long possible. In fact later altering or adjusting rules in production by moving classification thresholds can be very useful. Once one takes the care to work with scores, data re-weighting becomes much less attractive as most of the *perceived* benefit is now redundant to threshold adjustment.

What we are trying to say: there is a minor problem that classification rules are used too often, when numeric scores are usually much more useful. Obviously numeric scores must be translated into classification rules for many applications. However, this is best done at the point of application, not early in the modeling process. One noticeable distortion of switching to classification rules too early is the false impression that outcome proportional data re-weighting is more useful than it really is and needed more places than it really is.

### The Plan

In this note we will characterize some cases where such a data re-weighting or re-balancing is monotone. In these cases the re-weighting re-orders no predictions, so does nothing that a threshold change wouldn't do better. It is our hope that working a few examples will help demonstrate our thesis in a non-confrontational manner.

### The Non-Monotone Case

There are, of course, cases where re-weighting the data by the outcome class can change the order of predictions. For example, we demonstrate a non-trivial reordering effect with two variable (plus intercept) logistic regression [here](https://github.com/WinVector/Examples/blob/main/rebalance/rw_example_R.md). In this case we do get different models with different data weightings, but one still must determine which of these different models in in fact better for a given application.

## Single Numeric Variable (plus intercept) Logistic Regression

Let's look at an example where outcome proportional data re-weighting is provably no different than just choosing a different score threshold: single variable (plus intercept) logistic regression models. On the way we can prove a really fun lemma about the sign of one of the logistic regression coefficients.

Nina Zumel has a great article on proving theorems about logistic regression [here](https://win-vector.com/2011/09/14/the-simpler-derivation-of-logistic-regression/). For our purposes, a single variable (plus intercept) logistic regression is a probability model of the following form (examples in [`R` code)](https://www.r-project.org)).

```{r}
sigmoid <- function(x) {
  1/(1 + exp(-x))
}

predict_logistic <- function(
  a,  # scalar, model intercept
  b,  # scalar, model linear term
  x   # vector of instances of 
      # our single explanatory variable
  ) {
  sigmoid(a + b * x)
}
```

The above is what we mean by "a single variable (plus intercept) logistic regression model" (sometimes abbreviated as "a single variable logistic regression model"). The "single variable" is the unknown value `b`, and the "intercept" is `a`.

Typically we will have a training data set with has a numeric explanatory variable `x` and the dependent variable `y` taking values from `c(TRUE, FALSE)` that we are trying to predict. For example we might have data such as the following.

```{r}
d <- data.frame(
  x = c(-5,   0,     1),
  y = c(TRUE, FALSE, TRUE))

knitr::kable(d)
```

And could then fit a model as follows.

```{r}
m <- glm(
  y ~ x,
  data = d,
  family = binomial())

m$coefficients
```

Predictions can be performed using the standard `predict()` method.

```{r}
predict(m, newdata = d, type = 'response')
```

Or, using our example method.

```{r}
predict_logistic(
  a = m$coefficients['(Intercept)'],
  b = m$coefficients['x'],
  x = d$x)
```

### Re-Weighting Data Versus Moving The Classification Threshold

What we want to show is: re-weighting the training data proportional to the explanatory variable for a single variable model doesn't re-order predictions. This means for single-variable logistic models there is no point in such re-weighting of data, as the exact same set of classification rules can be found by moving the classification threshold. In more detail, altering `k` in the following can trade sensitivity/specificity or precision/recall.

```{r, eval = FALSE}
positive <- predict(
  glm(
    y ~ x,
    data = d,
    family = binomial(), weights = k + d$y),
  newdata = d,
  type = 'response') >= 0.5
```

In `Python` `sklearn`'s `.predict()` method the comparison of the model score to the threshold of `0.5` is often hard-coded. This feeds the mistaken belief that to change the classification rule you have to change the score by a transform of the form `weights = k + d$y`. Our position is: prefer calling something like `.predict_proba()` over using `.predict()`.

It is much simpler to consider moving the threshold away from `0.5` before monkeying with the training and score (our notes on this can be found [here](https://win-vector.com/2020/08/07/dont-use-classification-rules-for-classification-problems/)). A note on exploring threshold trade-offs can be found [here](https://win-vector.com/2020/08/17/unrolling-the-roc/).

However, for single variable logistic models, the data re-weighting transform is monotone in the predictions. So the same family of classification rules can always be achieved by picking an appropriate threshold `t` in the following.

```{r, eval = FALSE}
positive <- predict(
  glm(
    y ~ x,
    data = d,
    family = binomial()),
  newdata = d,
  type = 'response') >= t
```

One advantage of this latest form is: it is easier to try many values of `t` than many values of `k`.

### The Sign Lemma

To show the monotone property we are gong to establish a really neat lemma: 

<blockquote>
The sign of `b = m$coefficients['x']` is the same the sign of `mean(d$x[d$y]) - mean(d$x[!d$y])`. 
</blockquote>

Or: the sign of the logistic regression coefficient is the same is the sign of the difference in conditonal means of the explanatory variable.

For our example data `mean(d$x[d$y]) - mean(d$x[!d$y])` is very easy to calculate, it is `r mean(d$x[d$y]) - mean(d$x[!d$y])`. So we know that `m$coefficients['x']` must be negative, without having to go to the full effort of estimating it. This also is exactly what we need to characterize single variable logistic regression and show in this case explanatory variable proportional data re-weighting is a mere monotone transform (though as [we mentioned earlier](https://github.com/WinVector/Examples/blob/main/rebalance/rw_example_R.md), this is not always the case when more variables are present).

It takes a little bit of effort to establish this lemma, as it is elementary so we tend to have to use small steps in the proof. However, this effort is to be preferred to claiming the observation is "obvious" before moving on.

To establish our lemma pick the value `o` such that `sum((1 - s(o)) y) = sum((s(o) (1 - y))`.  By the intermediate value theorem, there is such a value.  Suppose `(a, b)` is the solution to our logistic regression problem. Now pick `d = (o - a) / b`. We replace our explanatory variable vector `x` with the vector `x' = x + d`.

This replacement means the new solution to the logistic regression problem is `(a + b * d, b)`, or `(o, b)` (as `sigmoid(o + b x')` = `sigmoid(a + b x)`). The point is: shifting the data alters the intercept coefficient, but not the slope-coefficient `b`.  Also shifting the data does not alter `mean(d$x[d$y]) - mean(d$x[!d$y])`. As we are only interested in the relation between `sign(b)` and `sign(mean(d$x[d$y]) - mean(d$x[!d$y]))`, we can without loss of generality assume the problem is already in this form with the intercept coefficient equal to `o`. That is: if we could prove the result in for problems in this form, that would be enough to establish the result in general.

Logistic regression coefficients are the unique solution to a loss function called deviance (essentially a scaling of cross-entropy).  In our case the loss function is:

```{r, eval = FALSE}
loss <- function(b) {
  -2 * ( sum(y * log(sigmoid(o + b * x))) + 
           sum((1 - y) * log(1 - sigmoid(o + b * x))) )
}
```

It is a standard result that `loss()` is a one dimensional [convex function](https://en.wikipedia.org/wiki/Convex_function). So if we know the sign of the derivative of the loss at `b = 0`, then we know if the minimum is to the left or to the right of zero. That is: the derivative of the loss tells us the sign of the optimal `b`, what we are trying to establish.

The derivative with respect to `b` of `log(sigmoid(o + b x))` is `1-sigmoid(o + b x)`, and the derivative of `log(1-sigmoid(o + b x))` is `-sigmoid(o + b x)`. This means the derivative of the loss with respect to `b` evaluated at `b = 0` is:

```{r, eval = FALSE}
2 * ( sum(y * (1 - sigmoid(o)) * x) -
           sum((1 - y) * sigmoid(o) * x) )
```

By our choice of `o`, this is non-negative scaling of `mean(d$x[d$y]) - mean(d$x[!d$y])`. So have established our claim: `sign(mean(d$x[d$y]) - mean(d$x[!d$y])) = sign(m$coefficients['x'])`.

### Back To The Monotone Property

With our lemma shown, it is possible to see that explanatory variable proportional data re-weighting moves predicitons in a monotone fashion. Notice re-weighting all of the examples of the positive class by the same amount doesn't change `mean(d$x[d$y]) - mean(d$x[!d$y])`. Therefore the sign of `b = m$coefficients['x']` is not changed. As `sigmoid(a + b x)` is monotone in `x` for a known-sign `b` we now have: no predictions change order under this sort of data re-scaling.

And that is the result: re-weighting all of the positives by the same amount can't re-order predictions for a single variable (plus intercept) logistic regression model. Thus for this style model there is no point to the procedure, as from an prediction-order perspective it accomplishes nothing more than a threshold change. 

### Another Example: The Saturated Indicators Case

One case where re-weighting the data as a function of the outcome class is monotone (and therefore not very powerful) is the case of saturated indicators.

Suppose all of our explanatory variables only take on the values `0` and `1`. Further suppose we have so many explanatory variables that the system is saturated in the sense that:

<ul>
<li>If `xa` is an explanatory variables then `1 - xa` is also an explanatory variable.</li>
<li>If `xa` and `xb` are explanatory variables then `xa * xb` is either all zero, or already an explanatory variable.</li>
</ul>

Essentially we are adding enough interactions or derived variables that the sets corresponding to the indicators for an *algebra of sets* (family closed under union and complement).  This also means essentialy we have enough variables such that every subset of the partition of rows implied by the variables jointly has an indicator variable.

In this case the explanatory variables partition the training data into sets that are all identifiable by single variables. In this case it is simple to see logistic regression, with some slight regularization in place, becomes just a look-up table. Training examples associated with one of our disjoint partition sets are scored with a probability of `a / (a + b)`, where `a` is the number of positive examples for this set during training and `b` is the number of negative examples seen in this set during training (assuming we don't have a separated problem, i.e. that `a` and `b` are both greater that zero). Then up-weighting the positive examples replaces the prediction with `(a w) / (a w + b)`, where `w` is our positive re-weighting factor.  A little algebra is enough to show that this transform, while non-linear, is monotone.  If `a / (a + b) > c / (c + d)` then `(a w) / (a w + b) > (c w) / (c w + d)`.

So in the fully saturated indicators case, re-weighting the data as a function of the positive class is also monotone.

## Single Categorical Variable

Notice the saturated indicators has variables that partition the training data. This data partition can be realized a single categorical variable. In fact any such partition or single categorical variable model is going to have the predictions move in a monotone fashion when up-sampling data ([example](https://github.com/WinVector/Examples/blob/main/rebalance/rw_example_R.md)).

## Conclusion

We have some some examples where re-weighting the data as a function of the class to be predicted does nothing interesting, where it is a monotone transform.  In these cases the same set of classification rules that can be found by such a re-weighting scheme can all be found just by picking different classification rule thresholds. Picking classification rule thresholds is much easier than re-training models and is something that can even be used as a useful adjustment later in production.

In addition to the examples, I also think the sign lemma is of interest. It emphasizes that coefficient sign is about the relative position of the conditional means, which is very much how methods such as linear discriminant analysis treat data.


