---
title: "install"
author: "Win-Vector LLC"
date: "December 3, 2015"
output: html_document
---

Here are the packages we most commonly use in our courses.
Note you could change the repository to another mirror 
(such as  http://cran.rstudio.com ) and also try using https based
repositories.  If pressing "knit" on this worksheet doesn't work
try copying and pasting the contents into your R prompt 
to run (which will allow more interactive dialogues such
as where to write the libraries and mirror selection).

```{r install1}
install.packages(c('devtools', 'gdata', 'gbm', 'glmnet', 'ggplot2',
                   'kernlab', 'knitr', 'mclust', 'mixtools', 'mgcv',
                   'plyr', 'pROC', 'randomForest', 'reshape2', 'ROCR',
                   'RSQLite',
                   'rpart', 'rpart.plot', 'scales', 'sqldf', 'snow', 'vtreat'),
                 repos='http://cran.us.r-project.org')
```

This next part can fail as devtools uses the RCurl package
for the downloads, and this can fail if the host machine doesn't
have a working libcurl installed for RCurl to use.

```{r optional}
tryCatch(
  devtools::install_github('WinVector/WVPlots'),
  error=function(e) print(e)
)
```


