---
title: "CodingExampleR"
author: "Win-Vector LLC"
date: "December 18, 2015"
output: html_document
---

```{r, echo=FALSE, results='hide', warning=FALSE}
source('utils.R')
source('modelR.R')
source('mkExample.R')
sourcedFns <- ls()
source('bindValues.R')
# devtools::install_github("WinVector/WVPlots")
library("WVPlots")
# devtools::install_github('WinVector/vtreat')
library('vtreat')

runAll <- TRUE
debug <- FALSE
cl <- NULL

set.seed(232567)
vplan <- list(designVar('x1',10),
              designVar('x2',10),
              designVar('x3',10),
              designVar('x4',10),
              designVar('x5',10),
              designVar('x6',10),
              designVar('x7',10),
              designVar('x8',10),
              designVar('x9',10),
              designVar('x10',10),
              designNoiseVar('n1',500),
              designNoiseVar('n2',500),
              designNoiseVar('n3',500),
              designNoiseVar('n4',500),
              designNoiseVar('n5',500),
              designNoiseVar('n6',500),
              designNoiseVar('n7',500),
              designNoiseVar('n8',500),
              designNoiseVar('n9',500),
              designNoiseVar('n10',500),
              designNoiseVar('n11',500),
              designNoiseVar('n12',500),
              designNoiseVar('n13',500),
              designNoiseVar('n14',500),
              designNoiseVar('n15',500),
              designNoiseVar('n16',500),
              designNoiseVar('n17',500),
              designNoiseVar('n18',500),
              designNoiseVar('n19',500),
              designNoiseVar('n20',500),
              designNoiseVar('n21',500),
              designNoiseVar('n22',500),
              designNoiseVar('n23',500),
              designNoiseVar('n24',500),
              designNoiseVar('n25',500),
              designNoiseVar('n26',500),
              designNoiseVar('n27',500),
              designNoiseVar('n28',500),
              designNoiseVar('n29',500),
              designNoiseVar('n30',500))
yVars <- c('yCat','yNumeric')
yName <- 'yNumeric'

dTrain <- generateExample(vplan,2000)  # Training set
vars <- setdiff(colnames(dTrain),yVars)
dCal <- generateExample(vplan,10000)   # Used to pick sigma
dTest <- generateExample(vplan,10000)  # Pure holdout test
```

Part of the idea is that for the Laplace noising to work we have
to plug in a sigma (level of noising).  We simulate having a very
good methodology to do so by supplying dCal a large calibration set
to pick sigma.  In practice you don't have such a set and would need
to either know sigma from first principles or experience, or use some
of your training data to build it.  What we want to demonstrate
is the effectiveness of the differential privacy inspired 
Laplace nosing technique, so we will give it a good sigma (which one
may or may not have in actual practice).


```{r, echo=FALSE, results='hide', warning=FALSE}
cl <- NULL
if(runAll && (!debug)) {
  cl <- parallel::makeCluster(parallel::detectCores())
}
```

```{r, echo=FALSE, results='hide', warning=FALSE}
mkWorker1 <- function() {
  bindToEnv(objNames=sourcedFns,
            yName,
            yVars,
            dTrain,
            vars,
            dCal)
  function(sigma) {
    bCoder <- trainEffectCoderR(dTrain,yName,vars,sigma)
    dTrainB <- bCoder$codeFrameR(dTrain)
    dCalB <- bCoder$codeFrameR(dCal)
    varsB <- setdiff(colnames(dTrainB),yVars)
    formulaB <- paste(yName,paste(varsB,collapse=' + '),sep=' ~ ')
    modelB <- lm(formulaB,data=dTrainB)
    dCalB$pred <- predict(modelB,newdata=dCalB)
    scoreB <- rmse(dCalB$pred,dCalB[[yName]])
    list(scoreB=scoreB,sigma=sigma)
  }
}

bSigmaBest = 0

if(runAll) {
  sigmaTargets <- (seq_len(41)-1)
  worker <- mkWorker1()
  if(!is.null(cl)) {
    results <- parallel::parLapplyLB(cl,sigmaTargets,worker)
  } else {
    results <- vector(mode='list',length=length(sigmaTargets))
    for(ii in seq_len(length(sigmaTargets))) {
      results[[ii]] <- worker(sigmaTargets[[ii]])
    }
  }
  
  bestB = Inf
  for(res in results) {
    sigma <- res$sigma
    scoreB <- res$scoreB
    if(scoreB<bestB) {
      bestB <- scoreB
      bSigmaBest <- sigma
    }
  }
}

print(paste('bSigmaBest',bSigmaBest))
```



```{r}
print('naive effects model')
bCoder <- trainEffectCoderR(dTrain,yName,vars,0)
dTrainB <- bCoder$codeFrameR(dTrain)
dTestB <- bCoder$codeFrameR(dTest)
varsB <- setdiff(colnames(dTrainB),yVars)
formulaB <- paste(yName,paste(varsB,collapse=' + '),sep=' ~ ')
modelB <- lm(formulaB,data=dTrainB)
print(summary(modelB))
dTrainB$pred <- predict(modelB,newdata=dTrainB)
print(paste('train rmse',rmse(dTrainB$pred,dTrainB[[yName]])))
print(WVPlots::ScatterHist(dTrainB,'pred',yName,
                           'naive effects model train',
                           smoothmethod='lm',annot_size=2))
dTestB$pred <- predict(modelB,newdata=dTestB)
print(paste('test rmse',rmse(dTestB$pred,dTestB[[yName]])))
print(WVPlots::ScatterHist(dTestB,'pred',yName,
                           'naive effects model test',
                           smoothmethod='lm',annot_size=2))
# try re-fitting pred to see if it is just a shift/scale issue
# it is unlikely to be such, but good to look
f2 <- paste(yName,'pred',sep=' ~ ')
m2 <- lm(f2,data=dTestB)  # dileberately fitting on test itself (to show it does not help)
print(summary(m2))
dTestB$pred2 <- predict(m2,newdata=dTestB)
print(paste('adjusted test rmse',rmse(dTestB$pred2,dTestB[[yName]])))
print(WVPlots::ScatterHist(dTestB,'pred2',yName,
                           'naive effects model test (adjusted on test)',
                           smoothmethod='lm',annot_size=2))
```


```{r}
print(paste('effects model, sigma=',bSigmaBest))
bCoder <- trainEffectCoderR(dTrain,yName,vars,bSigmaBest)
dTrainB <- bCoder$codeFrameR(dTrain)
dTestB <- bCoder$codeFrameR(dTest)
varsB <- setdiff(colnames(dTrainB),yVars)
formulaB <- paste(yName,paste(varsB,collapse=' + '),sep=' ~ ')
modelB <- lm(formulaB,data=dTrainB)
print(summary(modelB))
dTrainB$pred <- predict(modelB,newdata=dTrainB)
print(paste('train rmse',rmse(dTrainB$pred,dTrainB[[yName]])))
print(WVPlots::ScatterHist(dTrainB,'pred',yName,
                           paste('effects model train, sigma=',bSigmaBest),
                           smoothmethod='lm',annot_size=2))
dTestB$pred <- predict(modelB,newdata=dTestB)
print(paste('test rmse',rmse(dTestB$pred,dTestB[[yName]])))
print(WVPlots::ScatterHist(dTestB,'pred',yName,
                           paste('effects model test, sigma=',bSigmaBest),
                           smoothmethod='lm',annot_size=2))
```

```{r}
print('effects model, jacknifed')
bCoder <- trainEffectCoderR(dTrain,yName,vars,0)
# dTrainB <- bCoder$codeFrame(dTrain)
# dTrainB <- bCoder$codeFrame(dCal)
dTrainB <- jackknifeEffectCodeR(dTrain,yName,vars)
dTestB <- bCoder$codeFrameR(dTest)
varsB <- setdiff(colnames(dTrainB),yVars)
formulaB <- paste(yName,paste(varsB,collapse=' + '),sep=' ~ ')
modelB <- lm(formulaB,data=dTrainB)
print(summary(modelB))
dTrainB$pred <- predict(modelB,newdata=dTrainB)
print(paste('train rmse',rmse(dTrainB$pred,dTrainB[[yName]])))
print(WVPlots::ScatterHist(dTrainB,'pred',yName,
                       'effects model train, jackknifed',
                         smoothmethod='lm',annot_size=2))
dTestB$pred <- predict(modelB,newdata=dTestB)
print(paste('test rmse',rmse(dTestB$pred,dTestB[[yName]])))
print(WVPlots::ScatterHist(dTestB,'pred',yName,
                       'effects model test, jackknifed',
                         smoothmethod='lm',annot_size=2))
```


```{r}
print("vtreat split model")
pruneSig = 0.05

print("working vtreat split model")
mTitle <- 'vtreat split model'
isTrain <- runif(nrow(dTrain))<=0.5
dTrainDT <- dTrain[isTrain,]
dTrainDC <- dTrain[!isTrain,]
treatments <- vtreat::designTreatmentsN(dTrainDC,vars,yName,
                                        rareSig=0.3,
                                        smFactor=5.0,
                                        minFraction=2.0,
                                        verbose=FALSE,
                                        parallelCluster=cl)
dTrainV <- vtreat::prepare(treatments,dTrainDT,pruneSig=pruneSig,
                           parallelCluster=cl)

#print(treatments$scoreFrame)
varsV <- intersect(colnames(dTrainV),
                   treatments$scoreFrame$varName[treatments$scoreFrame$sig<pruneSig])
print(varsV)
dTestV <- vtreat::prepare(treatments,dTest,pruneSig=pruneSig,
                          varRestriction=varsV,
                          parallelCluster=cl)
formulaV <- paste(yName,paste(varsV,collapse=' + '),sep=' ~ ')
modelV <- lm(formulaV,data=dTrainV)
dTrainV$pred <- predict(modelV,newdata=dTrainV)
print(paste('train rmse',rmse(dTrainV$pred,dTrainV[[yName]])))
print(WVPlots::ScatterHist(dTrainV,'pred',yName,
                         paste(mTitle,'train'),
                         smoothmethod='lm',annot_size=2))
dTestV$pred <- predict(modelV,newdata=dTestV)
print(paste('test rmse',rmse(dTestV$pred,dTestV[[yName]])))
print(WVPlots::ScatterHist(dTestV,'pred',yName,
                         paste(mTitle,'test'),
                         smoothmethod='lm',annot_size=2))
```

```{r}
print("vtreat cross model")
pruneSig = 0.05
if("mkCrossFrameNExperiment" %in% ls('package:vtreat')) {
  print("working vtreat cross model")
  mTitle <- 'vtreat cross model'
  crossD <- vtreat::mkCrossFrameNExperiment(dTrain,vars,yName,
                                            rareSig=0.3,
                                            smFactor=5.0,
                                            minFraction=2.0,
                                            parallelCluster=cl)
  treatments <- crossD$treatments
  dTrainV <- crossD$crossFrame
  
#  print(treatments$scoreFrame)
  varsV <- intersect(colnames(dTrainV),
                     treatments$scoreFrame$varName[treatments$scoreFrame$sig<pruneSig])
  print(varsV)
  dTestV <- vtreat::prepare(treatments,dTest,pruneSig=pruneSig,
                            varRestriction=varsV,
                            parallelCluster=cl)
  formulaV <- paste(yName,paste(varsV,collapse=' + '),sep=' ~ ')
  modelV <- lm(formulaV,data=dTrainV)
  print(summary(modelV))
  dTrainV$pred <- predict(modelV,newdata=dTrainV)
  print(paste('train rmse',rmse(dTrainV$pred,dTrainV[[yName]])))
  print(WVPlots::ScatterHist(dTrainV,'pred',yName,
                             paste(mTitle,'train'),
                             smoothmethod='lm',annot_size=2))
  dTestV$pred <- predict(modelV,newdata=dTestV)
  print(paste('test rmse',rmse(dTestV$pred,dTestV[[yName]])))
  print(WVPlots::ScatterHist(dTestV,'pred',yName,
                             paste(mTitle,'test'),
                             smoothmethod='lm',annot_size=2))
} else {
  print("cross model not in library")
}
```


```{r}
sSigma = 100
print(paste('effects model, smoothing=',sSigma))
bCoder <- trainEffectCoderLSmooth(dTrain,yName,vars,sSigma)
dTrainB <- bCoder$codeFrameR(dTrain)
dTestB <- bCoder$codeFrameR(dTest)
varsB <- setdiff(colnames(dTrainB),yVars)
formulaB <- paste(yName,paste(varsB,collapse=' + '),sep=' ~ ')
modelB <- lm(formulaB,data=dTrainB)
print(summary(modelB))
dTrainB$pred <- predict(modelB,newdata=dTrainB)
print(paste('train rmse',rmse(dTrainB$pred,dTrainB[[yName]])))
print(WVPlots::ScatterHist(dTrainB,'pred',yName,
                           paste('effects model train, smoothing=',sSigma),
                           smoothmethod='lm',annot_size=2))
dTestB$pred <- predict(modelB,newdata=dTestB)
print(paste('test rmse',rmse(dTestB$pred,dTestB[[yName]])))
print(WVPlots::ScatterHist(dTestB,'pred',yName,
                           paste('effects model test, smoothing=',sSigma),
                           smoothmethod='lm',annot_size=2))
```




```{r}
if(!is.null(cl)) {
  parallel::stopCluster(cl)
  cl <- NULL
}
```
