---
title: "Linear Model Transfer"
author: "Nina Zumel"
date: "2024-07-08"
output: github_document
---

## Quantile Regression

In this notebook, we'll show an example of fitting quantile regressions with the `quantreg` package, 
then transferring the model to another linear framework: in this example `lm`. A similar procedure can be used 
to transfer an R `quantreg` model to a framework in another language, for example `scikit-learn` in Python, or `linfa` in Rust.

```{r}
library(quantreg)
library(ggplot2)
```

We'll use the Mammals dataset from the `quantreg` package for our example, and explore the relationship between mammal weight and mammal speed.

```{r}

data(Mammals)
summary(Mammals)

# let's compare log weight to log speed
Mammals = within(Mammals, 
                 {
                   logwt = log(weight)
                   logspeed = log(speed)
                 })

# hold out a few rows for didactic reasons
mammals = Mammals[1:100, ]
newmammals = Mammals[101:107, ]

```

We'll use `rq()` to fit models for the 10th and 90th percentile of log speed as a function of log weight, and whether or not the animal is a hopper (like a kangaroo) or "special": animals with "lifestyles in which speed does not figure as an important factor". Examples of "special" animals include the porcupine, the skunk, and the raccoon.

Note that, as discussed [here](http://www.econ.uiuc.edu/~roger/research/rq/FAQ), calculated percentile predictions aren't necessarily unique, but they are correct. Hence you may get a "non-uniqueness" warning when running `rq`.


```{r}
# 90th percentile
model09 = rq(logspeed ~ logwt + hoppers + specials, data=mammals, tau=0.9)
summary(model09)
coef(model09)

# 10th percentile
model01 = rq(logspeed ~ logwt + hoppers + specials, data=mammals, tau=0.1)
summary(model01)
coef(model01)

```

For comparison, we'll fit a linear model for expected log speed as a function of log weight, etc. This is mostly just to hammer in that the models we fit above with `rq` aren't the same.

```{r}
lmmodel = lm(logspeed ~ logwt + hoppers + specials, data=mammals)
summary(lmmodel)
coef(lmmodel)

```

Now let's compare actual log speeds to the linear and percentile predictions. We'll plot outcomes and predictions as a function of weight, with the data partitioned by the values of `specials` and `hoppers`.

```{r}
mammals$predmean = predict(lmmodel, newdata=mammals)
mammals$pred01 = predict(model01, newdata=mammals)
mammals$pred09 = predict(model09, newdata=mammals)

ggplot(mammals, aes(x=logwt, y=logspeed)) + 
  geom_point(color='darkgray') + 
  geom_line(aes(y=predmean), linetype="dashed") + 
  geom_line(aes(y=pred01), color='darkblue') +
  geom_line(aes(y=pred09), color='darkblue') + 
  facet_grid(hoppers ~ specials, labeller=label_both) + 
  ggtitle("Mammal speed as a function of weight, with 10th and 90th percentiles")


```


## Transfer the quantile model to the `lm` framework

Now let's transfer the coefficients from the `qr` models to `lm` models. As discussed in our main article (LINK TODO), all we need to do is evaluate our quantile models on a full rank set of rows. Since all the variables here are numeric or logical (0/1), we only need 4 independent rows: one for each variable, plus the intercept. 

Note that in our representation below, the intercept column is implicit, so you can think of `evalframe` having an invisible intercept column `c(1, 1, 1, 1)`, making `evalframe` a 4 by 4 full rank matrix.

First, we'll transfer the 90th percentile model.

```{r}
meanlogwt = mean(mammals$logwt) # not strictly necessary, but nice to keep the value in range

# you can build this frame any old way, this just happens to be readable
evalframe = wrapr::build_frame(
   'logwt', 'hoppers', 'specials' |
         0,  0,        0         | # for intercept
 meanlogwt,  0,        0         |
         0,  1,        0         |
         0,  0,        1
)

# hoppers and specials need to be logicals
evalframe$hoppers = as.logical(evalframe$hoppers)
evalframe$specials = as.logical(evalframe$specials)

# now add the predictions from the 90th percentile model
evalframe$logspeed = predict(model09, newdata=evalframe)

print(evalframe)
```

In the code above, I used `wrapr::build_frame` to specify the dataframe in a row-wise fashion. This is of course completely optional; I just didn't want to transpose the frame in my head to write it down in the default column-wise manner.

Now, let's fit an `lm` model to the evaluation frame

```{r}
model09lm = lm(logspeed ~ logwt + hoppers + specials, data=evalframe)

# compare the coefficients
data.frame(
  ported_model09 = coef(model09lm),
  original_model09 = coef(model09)
)

```

The coefficients of the new `lm` model are the same as the original 90th percentile model!
It's easy to prove that the predicted values are exactly the same, even on new data.

```{r}
Mammals$pred09 = predict(model09, newdata=Mammals)
Mammals$pred09x = predict(model09lm, newdata=Mammals)

# mark the training and holdout data
Mammals$datatype = c(rep('training', 100), rep('holdout', 7))

palette = c(training='darkgray', holdout='#d95f02')
ggplot(Mammals, aes(x=pred09, y=pred09x, color=datatype)) +
  geom_point() +
  geom_abline(color='gray') + 
  scale_color_manual(values=palette) + 
  ggtitle("90th percentile: lm representation vs original model")

```

We'll finish off by also re-fitting the 10th percentile model, and replicating the mean and percentile plot we created above.

```{r}
# get the 10th percentile predictions

evalframe$logspeed = predict(model01, newdata=evalframe)
model01lm = lm(logspeed ~ logwt + hoppers + specials, data=evalframe)

# compare the coefficients
data.frame(
  ported_model01 = coef(model01lm),
  original_model01 = coef(model01)
)
```

```{r}
# recreate the plot (including the holdout data)

Mammals$predmean = predict(lmmodel, newdata=Mammals)
Mammals$pred01 = predict(model01lm, newdata=Mammals)
Mammals$pred09 = predict(model09lm, newdata=Mammals)

ggplot(Mammals, aes(x=logwt, y=logspeed)) + 
  geom_point(aes(color=datatype)) + 
  geom_line(aes(y=predmean), linetype="dashed") + 
  geom_line(aes(y=pred01), color='darkblue') +
  geom_line(aes(y=pred09), color='darkblue') + 
  scale_color_manual(values=palette) + 
  facet_grid(hoppers ~ specials, labeller=label_both) + 
  ggtitle("Mammal speed as a function of weight, with 10th and 90th percentiles",
          subtitle="Using models ported to lm from qr")

```
