---
title: "Mercer_Kernel"
output: github_document
---

John Mount
December 2020

Here are some new notes on what is and what is not a Mercer Kernel style function.

  * `abs(x dot y)` and `relu(x dot y)` are not in general Mercer Kernels [https://github.com/WinVector/Examples/blob/main/Mercer_kernel/Mercer_Kernel.md](https://github.com/WinVector/Examples/blob/main/Mercer_kernel/Mercer_Kernel.md).
  * How to check if a 3 by 3 matrix is positive (semi) definite [https://github.com/WinVector/Examples/blob/main/Mercer_kernel/Dets.ipynb](https://github.com/WinVector/Examples/blob/main/Mercer_kernel/Dets.ipynb).

I've written before on Kernel methods (in the context of support vector machines).
 
  * [https://win-vector.com/2011/10/07/kernel-methods-and-support-vector-machines-de-mystified/](https://win-vector.com/2011/10/07/kernel-methods-and-support-vector-machines-de-mystified/)
  * [https://win-vector.com/2015/02/14/how-sure-are-you-that-large-margin-implies-low-vc-dimension/](https://win-vector.com/2015/02/14/how-sure-are-you-that-large-margin-implies-low-vc-dimension/)

Here we will show an example of non-kernel functions.

Both `abs(x dot y)` and `relu(x dot y)` (`=max(0, x dot y)`) are not positive semi-definite Mercer kernels when `x, y` are 2-dimensional vectors or larger. This can be subtle as for scalars `abs(x * y)` = `abs(x) * abs(y)` is exactly the kind of decomposition used to establish kernel properties.

By Mercer's theorem anything that generates only positive semi-definite Gram matrices can be realized as the inner product of vectors mapped into a, possibly infinite dimensional, vector space.

```{r}
n <- 4  # dimension of Gram matrix
d <- 2  # dimension of vectors

# abs(x.y) and relu(x.y) are not Mercer Kernels
#
# Gram matrix
# True for 3x3 case by Sylvester's criterion
# https://math.stackexchange.com/questions/1355088/is-the-absolute-value-of-a-p-d-s-matrix-p-d-s
# https://math.stackexchange.com/questions/1355088/is-the-absolute-value-of-a-p-d-s-matrix-p-d-s#comment2758302_1355088
# and true for scalar arguments as abs(x*y) = abs(x)*abs(y)

dot_f <- function(x, y) {
  sum(x*y)
}

relu_f <- function(x, y) {
  max(0, sum(x*y))
}

abs_f <- function(x, y) { 
  abs(sum(x*y))
}

# build the Gram matrix for a given kernel
Gram_matrix <- function(n, k_f, x) {
  idxs = expand.grid(
    i = seq_len(n), 
    j = seq_len(n))
  matrix(vapply(
    seq_len(nrow(idxs)), 
    function(i) k_f(x[idxs[i, 1], ], x[idxs[i, 2], ]), 
    numeric(1)),
    nrow = n,
    ncol = n)
}
```

```{r, eval=FALSE}
fn <- function(n, d) {
  # x <- matrix(rnorm(n = n*d), nrow = n, ncol = d)
  x <- matrix(
    sample(c(-1, 0, 1), 
           size = n*d, replace = TRUE), 
    nrow = n, 
    ncol = d)
  mat_relu <- Gram_matrix(n = n, k_f = relu_f, x = x)
  eigen_relu = min(eigen(mat_relu)$values)
  mat_abs <- Gram_matrix(n = n, k_f = abs_f, x = x)
  eigen_abs = min(eigen(mat_abs)$values)
  list(v = max(eigen_relu, eigen_abs), 
       x = x, 
       eigen_relu = eigen_relu, 
       eigen_abs = eigen_abs)
}

example <- lapply(
  1:10000,
  function(i) fn(n = n, d = d))

idx_relu = which.min(vapply(example, function(v) v$eigen_relu, numeric(1)))
print(example[[idx_relu]])

idx_abs = which.min(vapply(example, function(v) v$eigen_abs, numeric(1)))
print(example[[idx_abs]])

idx <- which.min(vapply(example, function(v) v$v, numeric(1)))
example <- example[[idx]]
print(example)
x <- example$x

t(x)
```

```{r}
x <- matrix(c(-1, -1, -1, 0, -1, 1, 0, -1), nrow = n, ncol = d)

t(x)
```

The dot-product is a Kernel, in fact the prototypical one. In fact `g^t G(a) g`, (`G(a)` being the Gram matrix with
`G(a)[i, j] = a(i) . a(j)`) is equal to `||[a_1, ... a_n] g||^2`.

```{r}
mat_dot <- Gram_matrix(n = n, k_f = dot_f, x = x)
mat_dot
```

```{r}
eigen_dot <- eigen(mat_dot)
if(min(eigen_dot$values)<0) {
  stop("expected eigen_dot to have only non-negative eigenvalues")
}
eigen_dot
```

```{r}
mat_abs <- Gram_matrix(n = n, k_f = abs_f, x = x)
mat_abs
```

```{r}
eigen_abs <- eigen(mat_abs)
eigen_abs
```

```{r}
example_abs <- eigen_abs$vectors[, n]
example_abs
```

```{r}
v_abs <- as.numeric(t(example_abs) %*% mat_abs %*% example_abs)
if(v_abs>=0) {
  stop("expected v_abs negative")
}
v_abs
```

```{r}
mat_relu <- Gram_matrix(n = n, k_f = relu_f, x = x)
mat_relu
```

```{r}
eigen_relu <- eigen(mat_relu)
eigen_relu
```

```{r}
example_relu <- eigen_relu$vectors[, n]
example_relu
```

```{r}
v_relu <- as.numeric(t(example_relu) %*% mat_relu %*% example_relu)
if(v_relu>=0) {
  stop("expected v_relu negative")
}
v_relu
```

