---
title: "Overcoming Omitted Variable Bias by Inferring the Hidden Data"
output: github_document
date: "2023-08-24"
always_allow_html: true
---

## Introduction

Let's continue along the lines discussed in [Omitted Variable Effects in Logistic Regression](https://win-vector.com/2023/08/18/omitted-variable-effects-in-logistic-regression/). 

The issue is as follows. For logistic regression, omitted variables cause estimation bias. This *even for independent variables* (which is not the case for more familiar linear regression).

This is a known problem with known mitigations:

  * John M. Neuhaus and Nicholas P. Jewell, (1993), "A Geometric Approach to Assess Bias Due to Omitted Covariates in Generalized Linear Models", Biometrika, Vol. 80, No. 4 (Dec., 1993), pp. 807-815.
  * Rhian Daniel, Jingjing Zhang, Daniel Farewell, (2020) "Making apples from oranges: Comparing noncollapsible effect estimators and their standard errors after adjustment for different covariate sets", Biometrical Journal, DOI: 10.1002/bimj.201900297.
  * Zhang, Zhiwei (2009) "Estimating a Marginal Causal Odds Ratio Subject to Confounding", Communications in Statistics - Theory and Methods, 38:3, 309 — 321.
  
(Thank you, Tom Palmer and Robert Horton for references!)

For this note, let's work out how to directly try and overcome the noted omitted variable bias.

## Our Example

For an example let's set up a logistic regression on two explanatory variables `x1` and `x2` . For simplicity we will take the case where `x1` and `x2` take on the values `0` and `1`.

Our data is then keyed by the values of these explanatory variables and the dependent or outcome variable `y`. The keying looks like the following.

```{r}
# show row labels
detailed_names <- expand.grid(
  x1 = c('0', '1'), 
  x2 = c('0', '1'), 
  y = c(FALSE, TRUE),
  stringsAsFactors = FALSE)
detailed_names <- detailed_names[
  order(detailed_names$y, 
        detailed_names$x2, 
        detailed_names$x1),
  , 
  drop = FALSE]

knitr::kable(detailed_names)
```

### The Data

Let's specify the joint probability distribution of our two explanatory variables. We choose them as independent with the following expected values.

```{r}
# specify explanatory variable distribution 
pX1 = 0.3
pX2 = 0.8
```

Our data set can then be completely described by above explanatory variable distribution *and* the conditional probability of the dependent outcomes. For our logistic regression problem we set up our outcome conditioning as `P[y == TRUE] ~ sigmoid(cval + b1 * x1 + b2 * x2)`. Our example coefficients are as follows.

```{r}
# 0.5772
(cval <- -digamma(1))
```

```{r}
# 3.1415
(b1 <- pi)
```

```{r}
# 27.182
(b2 <- -3 * exp(1))
```

Please remember these coefficients in this order for later.

```{r}
# show constants in an order will see again
c(cval, b1, b2)
```

Using the methodology of [Replicating a Linear Model](https://win-vector.com/2019/07/03/replicating-a-linear-model/) we can build a data set that obeys the specified explanatory variable distribution and has specified outcome probabilities. This is just us building a data set matching an assumed known answer. Our data distribution is going to be determined by `pX1`, `pX2`, and `P[y == TRUE] ~ sigmoid(cval + b1 * x1 + b2 * x2)`. Our inference task is to recover the parameters `pX1`, `pX2`, `cval`, `b1`, and `b2` from data, even in the situation where observers have omitted variable issues.

The complete detailed data is generated as follows. The `proportion` column is what proportion of a data set drawn from this specified distribution matches the row keys `x1`, `x2`, `y`, or is the joint probability of a given row type.

```{r}
# assign an example outcome or dependent variable
detailed_frame <- detailed_names
detailed_frame$x1 <- as.numeric(detailed_frame$x1)
detailed_frame$x2 <- as.numeric(detailed_frame$x2)
# get joint distribution of explanatory variables
detailed_frame$x_distribution <- (
  (detailed_frame$x1 * pX1 + (1 - detailed_frame$x1) * (1 - pX1))
    * (detailed_frame$x2 * pX2 + (1 - detailed_frame$x2) * (1 - pX2))
)
# get conditional probability of observed outcome
y_linear <- cval + b1 * detailed_frame$x1 + b2 * detailed_frame$x2
# converting "links" to probabilities
sigmoid <- function(x) {1 / (1 + exp(-x))}
y_probability <- sigmoid(y_linear)
detailed_frame$p_observed_outcome <- ifelse(
  detailed_frame$y, 
  y_probability, 
  1 - y_probability)
# compute joint explanatory plus outcome probability of each row
detailed_frame$proportion <- (
  detailed_frame$x_distribution * detailed_frame$p_observed_outcome)
stopifnot(  # abort render if this claim is not true
  abs(sum(detailed_frame$proportion) - 1) < 1e-6)
```

```{r}
knitr::kable(detailed_frame)
```

For a logistic regression problem, the relation between `x1`, `x2` and `y` is encoded in the `proportion` distribution that gives the joint expected frequency of each possible data row in a drawn sample.

```{r}
# clear some columns we are no longer using
detailed_frame$x_distribution <- NULL
detailed_frame$p_observed_outcome <- NULL
```

### Inferring From Fully Observed Data

We can confirm this data set encodes the expected logistic relationship by recovering the coefficients through fitting.

```{r}
# suppressWarnings() to avoid fractional data weight complaint
correct_coef <- suppressWarnings(
  glm(
    y ~ x1 + x2,
    data = detailed_frame,
    weights = detailed_frame$proportion,
    family = binomial()
  )$coef
)

correct_coef
```

Notice we recover the `cval + b1 * detailed_frame$x1 + b2 * detailed_frame$x2` form.

## The Problem

Now let's get to our issue. Suppose we have two experimenters, each of which only observing one of the explanatory variables. As we saw in [Omitted Variable Effects in Logistic Regression](https://win-vector.com/2023/08/18/omitted-variable-effects-in-logistic-regression/) each of these experimenters will in fact estimate coefficients that are **biased towards zero**, due to the non-collapsibility of the modeling set up. This differs from linear regression where for independent explanatory variables (as we have here) we would expect each experimenter to be able to get an unbiased estimate of the coefficient for the explanatory variable available to them!

To show this let's build a linear operator that computes the margins the experimenters actually observe.


```{r}
# build transfer matrix from joint observations to marginal observations
asterisk_symbol <- "&ast;"
margin_names <- rbind(
  expand.grid(
    x1 = c('0', '1'), 
    x2 = asterisk_symbol, 
    y = c(FALSE, TRUE), 
    stringsAsFactors = FALSE),
  expand.grid(
    x1 = asterisk_symbol, 
    x2 = c('0', '1'), 
    y = c(FALSE, TRUE), 
    stringsAsFactors = FALSE),
  expand.grid(
    x1 = c('0', '1'), 
    x2 = c('0', '1'), 
    y = asterisk_symbol, 
    stringsAsFactors = FALSE)
)
margin_transform <- matrix(
  data=0, 
  nrow = nrow(margin_names), 
  ncol = nrow(detailed_names))
colnames(margin_transform) <- paste0(
  "p(", 
  apply(detailed_names, 1, paste, collapse = ","), 
  ")")
rownames(margin_transform) <- paste0(
  "p(", 
  apply(margin_names, 1, paste, collapse = ","), 
  ")")
for (row_index in seq(nrow(margin_transform))) {
  for (col_index in seq(ncol(margin_transform))) {
    if (sum(detailed_names[col_index, ] == margin_names[row_index, ]) == 2) {
      margin_transform[row_index, col_index] = 1
    }
  }
}

knitr::kable(margin_transform, format = "html") |>
  kableExtra::kable_styling(font_size = 10)
```


```{r}
# apply the linear operator to compute marginalized observations
margin_frame <- margin_names
margin_frame$proportion <- as.numeric(
  margin_transform %*% detailed_frame$proportion)
rownames(margin_frame) <- paste0(
  "p(", 
  apply(margin_frame[, c('x1', 'x2', 'y')], 1, paste, collapse = ","), 
  ")")

knitr::kable(margin_frame)
```

The above margin frame describes how the detailed experiment is marginalized or censored down to different experimenters see. In our set-up experimenter 1 sees only the first four rows, and experimenter 2 sees only the next 4 rows. 

We consider the original detailed data as "unobserved", and parts of the marginalized results as what is observed by two experimenters.

### The Solution Strategy

We will see how collaborating experimenters, who never had a chance to examine the full detailed data, can estimate `proportion` and then estimate its pre-image under the `margin_transform`. This pre image is a complete distribution or description of the original (unobserved) detailed data. Once we have estimated or imputed the (previously) unobserved data we can then apply direct modeling techniques on that to complete the joint coefficient inference.

### Experimenter 1's view

Experimenter 1 only knows about the rows of the form `p(0/1, *, FALSE/TRUE)`. Experimenter 2 only knows about the rows named `p(*, 0/1, FALSE/TRUE)`. Neither experimenter knows the `p(0/1, 0/1, *)` rows. If they pool their data, they can in fact use the independence assumption to recover these rows.

Let's see what happens when each experimenter tries to perform inference on their fraction of the data.

```{r}
# select data available to d1
d1 <- margin_frame[
  margin_frame$x2 == asterisk_symbol, , drop = FALSE]
d1$x1 <- as.numeric(d1$x1)
d1$y <- as.logical(d1$y)

knitr::kable(d1)
```

```{r}
# solve from d1's point of view
suppressWarnings(
  glm(
    y ~ x1,
    data = d1,
    weights = d1$proportion,
    family = binomial()
  )$coef
)
```

Notice experimenter 1 got a *much* too small extimate of the `x1` coefficient. From experimenter 1's point of view, the effect of the omitted variable `x2` is making `x1` hard to correctly infer.

### Experimenter 2's view

```{r}
# select data available to d2
d2 <- margin_frame[
  margin_frame$x1 == asterisk_symbol, , drop = FALSE]
d2$x2 <- as.numeric(d2$x2)
d2$y <- as.logical(d2$y)

knitr::kable(d2)
```


```{r}
# solve from d2's point of view
suppressWarnings(
  glm(
    y ~ x2,
    data = d2,
    weights = d2$proportion,
    family = binomial()
  )$coef
)
```

Experimenter 2 also gets a wrong estimate, but closer as the the omitted `x1` variable had less influence.

## The Question

From the original data set's point of view: both experimenters have wrong estimates of their respective coefficients. The question then is: if the experimenters pool their effort can they infer the correct coefficients?

From our point of view only `d1` plus `d2` are observable, even after the experimenters choose to collaborate. This is because we are assuming neither of them had access to the original data as each failed to measure one of the explanatory variables.

Each experimenter knows a lot about the data. They known the distribution of their explanatory variable, and even the joint distribution of their explanatory and the dependent and outcome data. Assuming the two explanatory variables are independent, they even know the joint distribution of the explanatory variables. Together they *almost* know the original non-marginalized data distribution. Let's try to solve for an estimate of that distribution given the data available when the experimenters pool their observations.

This isn't the first time we have proposed a "guess at the original data, as it wasn't shared" as we played with this in [Checking claims in published statistics papers](https://win-vector.com/2013/04/08/checking-claims-in-published-statistics-papers/). It is a continual source of frustration that researchers are allowed to claim "results" without the backing code and data.

## A Solution

What we are going to do is: simulate experimenter 1 and experimenter 2 working together to try and recover a joint estimate of the `x1` and `x2` coefficients. We are going to use their pooled information to guess at plausible pre-images that represent the unobserved joint data set that has simultaneous `x1` and `x2` observations. This is kind of cute: instead of trying to invert the estimate bias, we try and guess at original data wher we know how to perform an unbiased analysis.

We characterize all pre-images of the pooled marginal information. These are all of the form of a pre-image of the estimated proportions column plus any elements of the marginalization processes' null-space (i.e. changes in data that are not respected by the summation process).

### The `x1` and `x2` distribution

Neither experimenter observed the following part of the marginal frame:

```{r}
# show x1 x2 distribution poriton of margin_frame
dx <- margin_frame[margin_frame$y == asterisk_symbol, , drop = FALSE]

knitr::kable(dx)
```

However, under their independence assumption they can estimate it from their observations as follows.

```{r}
# estimate x1 x2 distribution from d1 and d2
d1a <- aggregate(proportion ~ x1, data = d1, sum)
d2a <- aggregate(proportion ~ x2, data = d2, sum)
dxe <- merge(d1a, d2a, by = c())
dxe["proportion"] <- dxe$proportion.x * dxe$proportion.y
dxe$proportion.x <- NULL
dxe$proportion.y <- NULL
dxe <- dxe[order(dxe$x2, dxe$x1), , drop = FALSE]
stopifnot(  # abort render if this claim is not true
  all(dx[, c("x1", "x2")] == dxe[, c("x1", "x2")]))

knitr::kable(dxe)
```

Notice `dxe` is build only from `dx1` and `dx2` (plus the assumed independence of `x1` and `x2`). At this point we have inferred the `pX1` and `pX2` parameters from the observed data.

Putting this all together we get the joint experimenters estimate of the complete `proportion` vector.

```{r}
# put together experimenter 1 and 2's joint estimate of marginal proportions
# from data they have in their sub-experiments.
estimated_proportions <- c(
  d1$proportion,
  d2$proportion,
  dxe$proportion
)
stopifnot(  # abort render if this claim is not true
  max(abs(estimated_proportions - margin_frame$proportion)) < 1e-6)

estimated_proportions
```



### Solving For the Joint Distribution

In our case we have an example data distribution (with illegal negative entries!).

```{r}
# typical solution (in the linear sense, signs not enforced)
v <- as.numeric(
  MASS::ginv(margin_transform) %*% estimated_proportions)

v
```

And we have a single element of the null space, which is the direction different possible solutions vary in.

```{r}
# our degree of freedom between solutions
ns <- MASS::Null(t(margin_transform))
stopifnot(  # abort render if this claim is not true
  ncol(ns) == 1)
ns <- as.numeric(ns)

ns
```

So all valid solutions are of the form `v + z * ns` for scalars `z`. In fact all solutions are some interval of `z` values. Let's solve for them.

```{r}
# solve for the z's
proposed_solns <- unique(as.numeric(- v / ns )[abs(ns) > 1e-8])
solns <- proposed_solns[
  vapply(proposed_solns, 
         function(soln) {all(soln * ns + v >= -1e-8)}, 
         logical(1))]
solns <- unique(c(min(solns), max(solns)))
stopifnot(  # abort render if this claim is not true
  length(solns) > 0)
```

```{r}
# solve the logistic regression for each of our extreme guessed pre-images of the data
soln_i <- 0
soln_names <- rep('', length(solns))
for (soln in solns) {
  recovered <- as.numeric(soln * ns + v)
  recovered <- pmax(recovered, 0)  # get rid of any very near zero entries
  soln_i <- soln_i + 1
  soln_name <- paste0("recovered_", soln_i)
  detailed_frame[soln_name] <- recovered
  soln_names[soln_i] <- soln_name
}
```

Our attempted recovered solutions to the (unknown to either experimenter!) original data distribution details can be seen below.

```{r}
knitr::kable(detailed_frame)
```

As we can see these two extreme solutions are in fact actually fairly close. And the original (unobserved) data distribution is in fact a convex combination of these solutions.

```{r}
# confirm actual proportions in convex hull of solutions
convex_soln <- lm(
  p ~ 0 + r1 + r2, 
  data = data.frame(
    r1 = detailed_frame$recovered_1, 
    r2 = detailed_frame$recovered_2, 
    p = detailed_frame$proportion)
)
# abort render if these claims are not true
stopifnot(max(abs(convex_soln$residuals)) < 1e-8)
stopifnot(all(convex_soln$coefficients > -1e-8))
stopifnot(abs(sum(convex_soln$coefficients) - 1) < 1e-8)
```

### Inferring the logistic coefficients

And, in this case, it turns out each of our extreme solutions recovers essentially the same logistic regression solution! And this matches the unobserved detailed solution! Likely the one degree of freedom in the solution space matches some of the logistic regression balance conditions, and is therefore a unimportant or indifferent source of variation.

```{r}
# remind ourselves of the correct solution from actual (unobserved) joint data
correct_coef
```

```{r}
# run through our recovered solutions
soln_i <- 0
coef_list <- as.list(rep(NULL, length(soln_names)))
for (soln_name in soln_names) {
  print(soln_name)
  soln <- as.numeric(detailed_frame[[soln_name]])
  m_rec <- as.numeric(margin_transform %*% soln)
  max_error <- max(abs(m_rec - estimated_proportions))
  stopifnot(  # abort render if this claim is not true
    max_error < 1e-8)
  coef <- suppressWarnings(
    glm(
      y ~ x1 + x2,
      data = detailed_frame,
      weights = detailed_frame[[soln_name]],
      family = binomial()
    )$coef
  )
  stopifnot(  # abort render if this claim is not true
    max(abs(correct_coef - coef)) < 1e-6)
  print("recovered solution")
  print(coef)
  soln_i <- soln_i + 1
  coef_list[[soln_i]] <- coef
}
```

### Picking a point-estimate

It is convenient to pick a distinguished or "best guess" solution. In our case here it doesn't matter much, as both our extreme solutions can nearly identical logistic regression inferences. However if we do want to pick a single guess at the data pre-image the usual criterion is pick the maximum entropy one. 

This is just a simple principle: prefer flat distributions until one have evidence against them. This modeling technique is itself very strongly related to logistic regression modeling. Or one can say this is a bit opportunistic, we have an under-conditioned problem se we add an arbitrary convex criterion to pick a solution. Roughly we don't want to over-sell the maximum entropy pick, as we are already enforcing a lot of our sensible desiderata with the linear constraints.

```{r}
# brute force solve for maximum entropy mix
# obviously this can be done a bit slicker
entropy <- function(v) {
  v <- v[v > 0]
  v <- v / sum(v)
  -sum(v * log2(v))
}

opt_soln <- optimize(
  function(z) {
    entropy(
      z * detailed_frame$recovered_1 + 
        (1 - z) * detailed_frame$recovered_2)},
  c(0, 1),
  maximum = TRUE)

z_opt <- opt_soln$maximum
stopifnot(z_opt >= 0)  # abort render if this claim is not true
stopifnot(z_opt <= 1)  # abort render if this claim is not true
detailed_frame["maxent_dist"] <- (
  z_opt * detailed_frame$recovered_1 + 
    (1 - z_opt) * detailed_frame$recovered_2)
detailed_frame$recovered_1 <- NULL
detailed_frame$recovered_2 <- NULL
stopifnot(  # abort render if this claim is not true
  max(abs(detailed_frame$proportion - detailed_frame$maxent_dist)) < 1e-5)
```

Notice that the recovered `maxent_dist` is preternaturally close to the unobserved original `proportion`.

```{r}
knitr::kable(detailed_frame)
```

And we feel confident with the following estimated coefficients.

```{r}
recovered_coef <- suppressWarnings(
  glm(
    y ~ x1 + x2,
    data = detailed_frame,
    weights = detailed_frame[["maxent_dist"]],
    family = binomial()
  )$coef
)
stopifnot(  # abort render if this claim is not true
  max(abs(correct_coef - recovered_coef)) < 1e-6)

recovered_coef
```


## Conclusion

By pooling their observations the two researchers can recover a very good estimate of a joint analysis that neither of them performed. The strategy we used is: try to estimate plausible pre-images of the data that formed the observations that they saw, and then analyze that. This in fact gives us a method to invert the bias introduced by the omitted variables.

In the real world the two experimenters at best would be looking at marginalizations of different draws of related data. So we would not have exact matches we can invert- but instead would have to estimate low-discrepancy pre-images of the data. And, as always, as we are now introducing a lot of unobserved parameters we could go to Bayesian graphical model methods to sum this all out (instead of proposing a specific point-wise heuristic as we did here).
