---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

Our next "R and big data tip" is: summarizing big data.

We always say "if you are not looking at the data, you are not doing science"- and for big data you are very dependent on summaries (as you can't actually look at everything).

Simple question: is there an easy way to summarize big data in [`R`](https://www.r-project.org)?  

The answer is: yes, but we suggest you use the [`replyr`](https://CRAN.R-project.org/package=replyr) package to do so.

Let's set up a trivial example.

```{r setup}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
library("sparklyr")
packageVersion("sparklyr")
library("replyr")
packageVersion("replyr")

sc <- sparklyr::spark_connect(version='2.0.2', 
                              master = "local")
diris <- copy_to(sc, iris, 'diris')
```

The usual `S3`-`summary()` summarizes the handle, not the data.

```{r summary}
summary(diris)
```

`tibble::glimpse()` throws.

```{r glimpse, error=TRUE}
packageVersion("tibble")
# errors-out
glimpse(diris)
```

`broom::glance()` throws.

```{r glance, error=TRUE}
packageVersion("broom")
broom::glance(diris)
```

`replyr_summary()` works, and returns results in a `data.frame`.

```{r rsum}
replyr_summary(diris) %>%
  select(-nunique, -index, -nrows)
```


***


```{r cleanup}
sparklyr::spark_disconnect(sc)
rm(list=ls())
gc()
```
