---
title: "perm2"
author: "Win-Vector LLC"
date: "9/2/2017"
output: html_document
---

While deep in a large client project using [`Sparklyr`](https://spark.rstudio.com) and multinomial regression we recently ran into a problem:
[`Apache Spark`](https://spark.apache.org) chooses the order of multinomial regression outcome targets, whereas [`R`](https://www.r-project.org) users are 
used to choosing the order of the targets (please see [here](https://github.com/rstudio/sparklyr/issues/907) for some details).  So to make things more like `R` users expect, we need a way to translate one order to another.

Providing good solutions to gaps like this is one of the thing [Win-Vector LLC](http://www.win-vector.com/) does both in our [consulting](http://www.win-vector.com/site/service-offerings/consulting/) and [training](http://www.win-vector.com/site/service-offerings/training-overview/) practices.

Let's take a look at an example.  Suppose our two orderings are `o1` (the ordering `Spark ML` chooses) and `o2` (the order the `R` user chooses). 

```{r example}
set.seed(326346)
symbols <- letters[1:7]

o1 <- sample(symbols, length(symbols), replace = FALSE)
o1

o2 <- sample(symbols, length(symbols), replace = FALSE)
o2
```

To translate `Spark` results into `R` results we need a [permutation](https://en.wikipedia.org/wiki/Permutation) that takes `o1` to `o2`.  The idea is: if we had a permeation that takes `o1` to `o2` we could use it to re-map predictions that are in `o1` order to be predictions in `o2` order.

To solve this we crack open [our article on the algebra of permutations](http://www.win-vector.com/blog/2017/05/on-indexing-operators-and-composition/).  
 
We are going to use the fact that the `R` command `base::order(x)` builds a permutation `p` such that `x[p]` is in order.  

Given this the solution is: we find permutations `p1` and `p2` such that `o1[p1]` is ordered and `o2[p2]` is ordered.  Then build a permutation `perm` such that `o1[perm] = (o1[p1])[inverse_permutation(p2)]`.  I.e., to get from `o1` to `o2` move `o1` to sorted order and then move from the sorted order to `o2`'s order (by using the reverse of the process that sorts `o2`).  Again, the tools to solve this are in [our article on the relation between permutations and indexing](http://www.win-vector.com/blog/2017/05/on-indexing-operators-and-composition/).

Below is the complete solution (including combining the two steps into a single permutation):
 
```{r solution}
p1 <- order(o1)
p2 <- order(o2)

# invert p2
# see: http://www.win-vector.com/blog/2017/05/on-indexing-operators-and-composition/
p2inv <- seq_len(length(p2))
p2inv[p2] <- seq_len(length(p2))

(o1[p1])[p2inv]

# composition rule: (o1[p1])[p2inv] == o1[p1[p2inv]]
# see: http://www.win-vector.com/blog/2017/05/on-indexing-operators-and-composition/
perm <- p1[p2inv]
o1[perm]
```

The equivilence "`(o1[p1])[p2inv] == o1[p1[p2inv]]`" is frankly magic (though also quickly follows "by definition"), and studying it is the topic of [our original article on permutations](http://www.win-vector.com/blog/2017/05/on-indexing-operators-and-composition/).  

The above application is a good example of why it is nice to have a little theory worked out, even before you think you need it.



