---
title: "Followup: Adjusting Saturated Multivariate Linear Models"
author: "Nina Zumel"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r}
library(ggplot2) # for plotting
library(poorman) # for data wrangling
library(mgcv)    # for GAM
library(VGAM)    # for Tobit
library(lattice) # for levelplot/heatmap
library(akima)   # for interpolating a 3-d grid from irregular data
library(viridisLite) # for levelplot colors

set.seed(20240812)
```

This is a followup to the article [*Post-hoc Adjustment for Zero-Thresholded Linear Models*](https://win-vector.com/2024/08/16/post-hoc-adjustment-for-zero-thresholded-linear-models/), in which I
showed how to use a one-variable GAM (e.g., a spline) to adjust a linear model in a problem space 
where outcomes are strictly nonnegative. If you haven't read that article, I suggest you check it out, first.

> When you don’t expect to see too many zeros in practice, modeling the process as linear and thresholding negative predictions at zero is not unreasonable. But the more zeros (saturations) you expect to see, the less well a linear model will perform.

The original motivation for the work we did here was to help a client who had built a fairly complex multivariate model to predict count data. Their underlying assumption was that in their domain, a count of zero is a rare event. Their model training and deployment process was automated and put into production across multiple sites. 

Unfortunately, zero-counts are not so rare as they originally believed, at some of their sites. Because coming up with and deploying a new model design is not necessarily feasible at this stage (at least not in the immediate term), they wanted to figure out how to adjust their existing models to operate even at high zero-count sites. Note that this adjustment is a *one-dimensional* process: mapping the output of one model to a prediction that is closer to the actual outcomes.

To motivate this adjustment process in the previous article, I used an example of a linear model that was fit to a one-dimensional saturated process. This was so I could plot the resulting "hockey stick" function, and show what happens with different model adjustments. But it's been my experience that people won't believe that this procedure will generalize to multivariate linear models, unless I show them that it works on such models. So in this article, I'll apply the same procedures to a linear model that was fit to two variables, just to prove the point.

I'll also add an additional adjustment that wasn't in the original article: adjustment using a Tobit model. You can see more details of that in a revised version of the article on github, [here](https://github.com/WinVector/Examples/blob/main/linear_regression_w_zeros/lm_adjust_wtobit.md). Tobit adjustments work nearly as well as GAM adjustments, and do have the (potential) advantage of having a stronger inductive bias, if you believe that your process is truly linear in it's non-saturated regions [^1].

[^1]: Statisticians generally use the word "censored" when talking about processes that threshold out to a minimum or maximum value (or both). Often, the assumption is that the measurements are censored because of limitations of the measurement itself: for example, you can't track a subject for more than five years, so the longest lifetime you can expect to measure is 5, even if the subject lives for decades longer. So models like Tobit try to predict *as if* the data isn't censored: they can predict a lifetime longer than 5 years, even though those lifetimes can't be measured. I am using the word "saturated" to indicate that the outcome being measured literally cannot go beyond the threshold: counts can never be negative. A possibly useful analogy is the image from a digital camera: too much light "blows out" the photo. If you are trying to predict the intensity of the light that hit the camera sensor using the image pixels as the outcome data, then the data is censored. If you are trying to predict *the values of the pixels*, then the data is saturated.

In the second part of the article, I'll try fitting a model directly to the input data, using both GAM and Tobit, since ideally, that's what you'd do if you were approaching this problem *de novo*.

For legibility, I'm going to hide a lot of the code when generating this article, but you can find the original R Markdown source code on github as well, [here](https://github.com/WinVector/Examples/blob/main/linear_regression_w_zeros/lm2d_adjust.Rmd).

## Example Problem

Here are the characteristics of our example problem:

* The feature `u` positively correlates with outcome, while `v` negatively correlates.
* The outcome `y` is constrained to be nonnegative; in other words, it "saturates" at zero.
* Both features `u` and `v` are bounded between 0 and 1 uniformly.
* There is a moderate noise process on top of it.


```{r echo=TRUE}
trueprocess = function(N) {
  u = runif(N)
  v = runif(N)
  noise = 0.3 * rnorm(N) 
  y = pmax(0, 4 * u - 3 * v + noise)
  data.frame(u = u, v = v, y = y)
}

```

```{r}

# take a slice of the ideal function
# for a fixed u
ideal_uslice = function(ufixed, v) {
  pmax(0, 4*ufixed - 3*v)
}

# take a rough slice of the data
# around a fixed u
uslice = function(ufixed, processdata) {
  in_slice = abs(processdata$u - ufixed) < 0.05
  subset(processdata, in_slice)
}

rmse = function(y, ypred) {
  err = y - ypred
  sqrt(mean(err^2))
}

# we'll also calculate bias: the mean value of (y - ypred)
bias = function(y, ypred) {
  err = ypred - y
  b = mean(err)
  # let's round down to zero for small numbers
  b = ifelse(abs(b) < 1e-12, 0, b)
  b
}

```


```{r}
#
# We will also define the procedures for model adjustment and for adjusted prediction that we showed in the previous article; 
# but for legibility, I'll hide this part of the code.
#


#
# Rescale the slope of the original model's positive region
#

# outcome here is the name of the outcome column in the data frame
fit_scaler = function(initial_model, outcome, data) {
  ypred0 = predict(initial_model, newdata=data)
  df = data.frame(ypred0 = ypred0, y=data[[outcome]])
  
  # reduce the data to non-negative predictions
  df = subset(df, ypred0 > 0)
  
  # now fit the new model to the non-negative predictions
  smodel = lm(y ~ 0 + ypred0, data=df) 
  
  # return a list: initial model, adjustment model
  list(initial_model=initial_model, adjustment=smodel)
 
}

#
# Adjust the original model's positive region
# via another linear model
#

fit_linmod = function(initial_model, outcome, data) {
  ypred0 = predict(initial_model, newdata=data)
  df = data.frame(ypred0 = ypred0, y=data[[outcome]])
  
  # reduce the data to non-negative predictions
  df = subset(df, ypred0 > 0)
  
  # now fit the new model
  lpmodel = lm(y ~ ypred0, data=df) 
  
  # return a list: initial model, adjustment model
  list(initial_model=initial_model, adjustment=lpmodel)
 
}

#
# Adjust the original model's positive region
# via a spline, implemented as a GAM model
#

fit_gammod = function(initial_model, outcome, data) {
  ypred0 = predict(initial_model, newdata=data)
  df = data.frame(ypred0 = ypred0, y=data[[outcome]])
  
  # reduce the data to non-negative predictions
  df = subset(df, ypred0 > 0)
  
  # now fit the new model
  gammodel = gam(y ~ s(ypred0), data=df) 
  
  # return a list: initial model, adjustment model
  list(initial_model=initial_model, adjustment=gammodel)
 
}

#
# tobit adjustment wasn't in the original post, but it is here:
# https://github.com/WinVector/Examples/blob/main/linear_regression_w_zeros/lm_adjust_wtobit.md
#
fit_tobit = function(initial_model, outcome, data) {
  ypred0 = predict(initial_model, newdata=data)
  df = data.frame(ypred0 = ypred0, y=data[[outcome]])
  
  # reduce the data to non-negative predictions
  df = subset(df, ypred0 > 0)
  
  # now fit the new model
  tobitmodel = vglm(y ~ ypred0, tobit(Lower = 0), data = df)
  
  # return a list: initial model, adjustment model
  list(initial_model=initial_model, adjustment=tobitmodel)
}

#
# Make predictions using any of the above adjusted models
#

# "model" is a list (initial_model, adjustment)
do_predict = function(model, newdata) {
  mod0 = model$initial_model
  adjmod = model$adjustment
  
  df = data.frame(ypred0 = predict(mod0, newdata=newdata))
  # if the adjustment model makes negative predictions, threshold to 0
  if(class(adjmod)[1] =="vglm") { # the tobit model
    yadj = pmax(0, predict(adjmod, df)[, 'mu'])
  } else {
    yadj = pmax(0, predict(adjmod, newdata=df))
  }

  # if linear model predicts a negative number,
  # predict 0, else use adjusted model
  ypred = ifelse(df$ypred0  <= 0, 0, yadj)
  ypred
}




```

## The Data

Let's generate some training data.

```{r echo=TRUE}
traind = trueprocess(1000)
head(traind)
```

Let's plot a heatmap of the training data. Note that there's lots of saturation.

```{r}
# https://stackoverflow.com/questions/7536577/using-color-as-the-3rd-dimension
ak.interp = with(traind, interp(u, v, y, duplicate="mean")) # list (x, y, z)

palette <- viridis(100)
levelplot(ak.interp$z, main="True process", col.regions=palette, contour=FALSE)
                 
```

We're going to use all the data to train our models, but it's not easy to show comparison plots in three dimensions in a way that's legible. So we'll also look at a slice of the data, by holding `u` fixed to 0.5 (or close to 0.5, at least). 

```{r}
# the ideal process
ufixed = 0.5
v = seq(from=0, to=1, by=0.01)

# the ideal slice
idealf = data.frame(v = v, y = ideal_uslice(ufixed, v))
# a slice from the training data
slicef = uslice(ufixed, traind)

theme_set(theme_bw()) # set global ggplot theme

ggplot(mapping=aes(x=v, y=y)) + 
  geom_point(data=slicef, color="gray") +
  geom_line(data=idealf, linetype="dashed") + 
  ggtitle("Slice of training data around u=0.5", 
          subtitle="Dashed line is ideal process")
```

As you can see, the data is fairly noisy, in addition to being quite saturated. 
Now let's train a linear model on the data, as well as all our adjusted models.

## Part I: Adjustments to a Linear Model

### Train the initial model, and all the adjustments

We'll fit the model, make the naive adjustment of thresholding the data at zero,
then fit all the candidate adjustment models. Note that only the inital model ever
makes reference to the input variables.

```{r echo=TRUE}
# train the initial model, 
# get the initial predictions and thresholded predictions

initial_model = lm(y ~ u + v, data=traind)
traind$y_initial= predict(initial_model, newdata=traind)
traind$y_pred0 = pmax(0, traind$y_initial)

# fit the adjustment models
# (function definitions in R markdown document)
scaling_model = fit_scaler(initial_model, "y", traind)
linadj_model = fit_linmod(initial_model, "y", traind)
gamadj_model = fit_gammod(initial_model, "y", traind)
tobitadj_model = fit_tobit(initial_model, "y", traind)

adjustments = list(scaling_model, linadj_model, gamadj_model, tobitadj_model)
names(adjustments) = c("y_linscale", "y_linadj", "y_gamadj", "y_tobitadj")

# make the predictions
for(adj in names(adjustments)) {
  traind[[adj]] = do_predict(adjustments[[adj]], newdata=traind)
}
```

Let's look at the RMSE and bias of each model.

```{r}
# pivot to a form that's better for plotting and summarization
traindlong = pivot_longer(
  traind,
  c("y_initial", "y_pred0", "y_linscale", "y_linadj", "y_gamadj", "y_tobitadj"),
  names_to = "prediction_type",
  names_prefix = "y_",
  values_to = "prediction"
)

errframe = traindlong |>
  group_by(prediction_type) |>
  summarize(RMSE = rmse(y, prediction),
            bias = bias(y, prediction)) |>
  ungroup()
rownames(errframe) = errframe$prediction_type

# reorder, for presentation
# descriptions
descv = c("initial model", "zero-thresholded model", "scale-adjusted model", "linear-adjusted model", 
          "GAM-adjusted model", "Tobit-adjusted model")
names(descv) = c("initial", "pred0", "linscale", "linadj", "gamadj", "tobitadj")

# a little rearrangement for good presentation
rownames(errframe) = errframe$prediction_type
errframe$description = descv[errframe$prediction_type]
errframe = errframe[names(descv), c("prediction_type", "description", "RMSE", "bias")]
knitr::kable(errframe, caption="Model RMSE and bias on training data", row.names=FALSE)

```

As expected (if you've read the previous article), the GAM-adjusted model has the lowest training RMSE, and also lower bias than the other adjusted models. The Tobit-adjusted model has comparable RMSE, and slightly more bias.

We can try to visualize what's happening, by holding `u` constant at different values and looking at slices of the prediction surfaces. 
We'll just compare the original linear model, GAM-adjustment, and Tobit-adjustment.

```{r fig.width=8, fig.height=7}
uvals = c(0.1, 0.25, 0.5, 1)
v = seq(from=0, to=1, by=0.01)

# holding u and label fixed, return a slice
# need to add the ideal
make_slice = function(ufixed, vvec, label) {
  df = data.frame(label=label, u=ufixed, v=v)
  if (label=="initial") {
    df$ypred = predict(initial_model, newdata=df)
    return(df)
  }
  if (label=="pred0") {
    df$ypred = pmax(0, predict(initial_model, newdata=df))
    return(df)
  }
  model = adjustments[[paste0("y_",label)]]
  df$ypred = do_predict(model, newdata=df)
  df
}

# holding u fixed, return a dataframe of slices
make_uslices = function(u, vvec, labelsvec) {
  slices = lapply(labelsvec, 
                  function(label) {make_slice(u, vvec, label)})
  bind_rows(slices)
}

# let's just look at initial, gamadj, and tobitadj
models = c("initial", "gamadj", "tobitadj")
slices = lapply(uvals, function(u) {make_uslices(u, vvec, models)})
slices = bind_rows(slices)

# the ideal slices
ideal = lapply(uvals, function(u) {data.frame(u=u, v=v, y=ideal_uslice(u, v))})
ideal = bind_rows(ideal)

# we can also superimpose approximate slices of the training data
datums = lapply(uvals, function(u) {
  slice = uslice(u, traind)
  slice$u = u # I need this for plotting
  slice
  })
datums = bind_rows(datums)


palette = c(initial = "#BBBBBB", pred0 = "#AA3377", linscale = "#228833", 
            linadj = "#4477AA",  gamadj = "#EE6677", tobitadj = "#66CCEE")

ggplot() + 
  geom_point(data=datums, mapping=aes(x=v, y=y), color="gray", alpha=0.5) +
  geom_line(data=ideal, mapping=aes(x=v, y=y), linetype="dashed", color="black") + 
  geom_line(data=slices, aes(x=v, y = ypred, color=label)) + 
  facet_wrap(~u, ncol=2, scales="free_y", labeller=label_both) +
  scale_color_manual(breaks=names(palette), values=palette) + 
  ggtitle("A comparison of model adjustments across various slices of the prediction surface",
          subtitle="Ideal process as dashed line") + 
  theme(legend.position="bottom")



```


### Test on holdout

Let's compare the models on holdout data. For this example the holdout performances are similar to training.

```{r echo=TRUE}
testd = trueprocess(5000)

testd$y_initial= predict(initial_model, newdata=testd)
testd$y_pred0 = pmax(0, testd$y_initial)
for(adj in names(adjustments)) {
  testd[[adj]] = do_predict(adjustments[[adj]], newdata=testd)
}

```

```{r}
# pivot to a form that's better for plotting and summarization
testdlong = pivot_longer(
  testd,
  c("y_initial", "y_pred0", "y_linscale", "y_linadj", "y_gamadj", "y_tobitadj"),
  names_to = "prediction_type",
  names_prefix = "y_",
  values_to = "prediction"
)

errframe = testdlong |>
  group_by(prediction_type) |>
  summarize(RMSE = rmse(y, prediction),
            bias = bias(y, prediction)) |>
  ungroup()
rownames(errframe) = errframe$prediction_type

# reorder, for presentation
rownames(errframe) = errframe$prediction_type
errframe$description = descv[errframe$prediction_type]
errframe = errframe[names(descv), c("prediction_type", "description", "RMSE", "bias")]
knitr::kable(errframe, caption="Model RMSE and bias on training data", row.names=FALSE)
```


## Part II : Fitting directly to the input data

Now let's try both Tobit and GAM directly on the input data. There are, of course, other methods we could try, like Poisson or Zero-inflated Poisson regression. But since we've been focusing on Tobit and GAM as adjustments, we'll just stick to those.

Note that since neither of these models are restricted to positive predictions, we still have to threshold to zero.

Let's fit to our training data.
```{r echo=TRUE}

gam_model = gam(y ~ s(u) + s(v), data=traind)
traind$y_gam = pmax(0, predict(gam_model, newdata=traind))

tobit_model = vglm(y ~ u + v, tobit(Lower=0), data=traind)
# we still have to threshold the tobit model at 0
traind$y_tobit = pmax(0, predict(tobit_model, traind)[, "mu"])

```

We'll compare the full model fits to the GAM-adjusted and Tobit-adjusted linear models.

```{r}
# pivot to a form that's better for plotting and summarization
traindlong = pivot_longer(
  select(traind, u, v, y, y_gamadj, y_tobitadj, y_gam, y_tobit),
  c("y_gamadj", "y_tobitadj", "y_gam", "y_tobit"),
  names_to = "prediction_type",
  names_prefix = "y_",
  values_to = "prediction"
)

errframe = traindlong |>
  group_by(prediction_type) |>
  summarize(RMSE = rmse(y, prediction),
            bias = bias(y, prediction)) |>
  ungroup()
rownames(errframe) = errframe$prediction_type

# reorder, for presentation
# descriptions
descv = c("GAM-adjusted model", "Tobit-adjusted model", "Full GAM model", "Full Tobit model")
names(descv) = c("gamadj", "tobitadj",  "gam", "tobit")

# a little rearrangement for good presentation
rownames(errframe) = errframe$prediction_type
errframe$description = descv[errframe$prediction_type]
errframe = errframe[names(descv), c("prediction_type", "description", "RMSE", "bias")]
knitr::kable(errframe, caption="Model RMSE and bias on training data", row.names=FALSE)

```

The full (thresholded) Tobit model does slightly better than the Tobit-adjusted linear model, but the thresholded GAM doesn't do so well. Since this problem is truly linear, the stronger inductive bias of the Tobit model serves us well.

We can also evaluate these models on holdout data.


```{r}
testd$y_gam = pmax(0, predict(gam_model, newdata=testd))
testd$y_tobit = pmax(0, predict(tobit_model, testd)[, "mu"])

# pivot to a form that's better for plotting and summarization
testdlong = pivot_longer(
  select(testd, u, v, y, y_gamadj, y_tobitadj, y_gam, y_tobit),
  c("y_gamadj", "y_tobitadj", "y_gam", "y_tobit"),
  names_to = "prediction_type",
  names_prefix = "y_",
  values_to = "prediction"
)

errframe = testdlong |>
  group_by(prediction_type) |>
  summarize(RMSE = rmse(y, prediction),
            bias = bias(y, prediction)) |>
  ungroup()
rownames(errframe) = errframe$prediction_type

# reorder, for presentation
# descriptions
descv = c("GAM-adjusted model", "Tobit-adjusted model", "Full GAM model", "Full Tobit model")
names(descv) = c("gamadj", "tobitadj",  "gam", "tobit")

# a little rearrangement for good presentation
rownames(errframe) = errframe$prediction_type
errframe$description = descv[errframe$prediction_type]
errframe = errframe[names(descv), c("prediction_type", "description", "RMSE", "bias")]
knitr::kable(errframe, caption="Model RMSE and bias on training data", row.names=FALSE)
```

We get similar results. Let's plot some slices to get an idea of what's happening.

```{r}
# new make_slice functions
make_uslices2 = function(u, vvec, modellist) {
  slices = lapply(names(modellist),
                        function(label) {make_slice2(u, vvec, modellist[label])})
  bind_rows(slices)
}

# note onemodel is a length-1 list (so a labelled model)
make_slice2 = function(ufixed, vvec, onemodel) {
  label = names(onemodel)[1]
  model = onemodel[[1]]
  
  df = data.frame(label=label, u=ufixed, v=v)
  if(class(model)[1] == "vglm") {
    df$ypred = pmax(0, predict(model, df)[, 'mu'])
  } else {
    df$ypred = pmax(0, predict(model, newdata=df))
  }
  df
}
  
# let's just look at initial, gamadj, and tobitadj
modellist = list(initial_model, gam_model, tobit_model)
names(modellist) = c("initial", "gam", "tobit")
slices = lapply(uvals, function(u) {make_uslices2(u, vvec, modellist)})
slices = bind_rows(slices)

# the ideal slices
ideal = lapply(uvals, function(u) {data.frame(u=u, v=v, y=ideal_uslice(u, v))})
ideal = bind_rows(ideal)

# we can also superimpose approximate slices of the training data
datums = lapply(uvals, function(u) {
  slice = uslice(u, traind)
  slice$u = u # I need this for plotting
  slice
  })
datums = bind_rows(datums)


palette = c(initial = "#BBBBBB",  gam = "#EE6677", tobit = "#66CCEE")

ggplot() + 
  geom_point(data=datums, mapping=aes(x=v, y=y), color="gray", alpha=0.5) +
  geom_line(data=ideal, mapping=aes(x=v, y=y), linetype="dashed", color="black") + 
  geom_line(data=slices, aes(x=v, y = ypred, color=label)) + 
  facet_wrap(~u, ncol=2, scales="free_y", labeller=label_both) +
  scale_color_manual(breaks=names(palette), values=palette) + 
  ggtitle("A comparison of Thresholded Linear, GAM and Tobit models across various slices of the prediction surface",
          subtitle="Ideal process as dashed line") + 
  theme(legend.position="bottom")

```


