---
title: "Dependencies"
author: "Win-Vector LLC"
date: "12/1/2017"
output: github_document
---

This is an example of an erroneous calculation in `dplyr` `0.7.4` with databases, likely
arising from mishandling of expression to expression dependencies (which [we have written about
before](http://www.win-vector.com/blog/2017/09/my-advice-on-dplyrmutate/)).  Since this is
not a security issue, can cause non-signaled incorrect results, and has an easy fix: we have
decided to document it here and to [distribute an announcement](http://www.win-vector.com/blog/2017/12/please-inspect-your-dplyrdatabase-code/).

First we set up our example.

```{r setup}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")

my_db <- DBI::dbConnect(RSQLite::SQLite(),
                        ":memory:")
d <- dplyr::copy_to(
  my_db, 
  data.frame(valuesA = c("A", NA, NA),
             valuesB = c("B", NA, NA),
             canUseFix1 = c(TRUE, TRUE, FALSE),
             fix1 = c('Fix_1_V1', "Fix_1_V2", "Fix_1_V3"),
             canUseFix2 = c(FALSE, FALSE, TRUE),
             fix2 = c('Fix_2_V1', "Fix_2_V2", "Fix_2_V3"),
             stringsAsFactors = FALSE),
  'd', 
  temporary = TRUE, overwrite = TRUE)
knitr::kable(dplyr::collect(d))
```

For our example we are using `canUseFix1*` columns to find which positions
of our `values*` columns can be replaced by the corresponding fix values.
This is a common situation in data processing: where you have a column you
wish to populate from a ordered sequence of alternate sources.

We could write this as nested `ifelse()` or coalesce. But suppose 
we had written the code as below.

```{r badresult}
fixed <- dplyr::mutate(
  d,
  valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                    fix1, valuesA),
  valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                    fix2, valuesA),
  valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                    fix1, valuesB),
  valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                    fix2, valuesB))

fixed %>%
  dplyr::select(., valuesA, valuesB) %>%
  dplyr::collect(.) %>%
  knitr::kable(.)
```

Notice this *silently* failed! It gave a wrong answer, with no indicated error.

The third `valuesA` value remains at `NA` even though it should 
have been repaired by the fix 2 rule.  This is not due to order of statements as
the fix rules were deliberately chosen to apply to disjoint rows.

Looking further we see `dplyr` seem to generate incomplete `SQL` (not all the 
intended transforms seem to survive the translation, notice there are 3 `CASE WHEN`
statements in the generated `SQL`, not 4):

```{r badsql}
d  %>%
  dplyr::mutate(
    .,
    valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                      fix1, valuesA),
    valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                      fix2, valuesA),
    valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                      fix1, valuesB),
    valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                      fix2, valuesB)) %>%
  dplyr::show_query(.)
```

For our recommended current work-around, please see [here](http://winvector.github.io/FluidData/DplyrDependencies.html).

---------------


```{r seif}
sessionInfo()
```

Also note: as of `December 1, 2017` upgrading the development versions
of `dbplyr` and `dplyr` is *not* sufficient to fix the issue:

```r
devtools::install_github("tidyverse/dbplyr")
devtools::install_github("tidyverse/dplyr")

base::date()
#> [1] "Fri Dec  1 09:32:56 2017"
packageVersion("dbplyr")
#> [1] '1.1.0.9000'
packageVersion("dplyr")
#> [1] '0.7.4.9000'

my_db <- DBI::dbConnect(RSQLite::SQLite(),
                        ":memory:")
d <- dplyr::copy_to(
  my_db, 
  data.frame(valuesA = c("A", NA, NA),
             valuesB = c("B", NA, NA),
             canUseFix1 = c(TRUE, TRUE, FALSE),
             fix1 = c('Fix_1_V1', "Fix_1_V2", "Fix_1_V3"),
             canUseFix2 = c(FALSE, FALSE, TRUE),
             fix2 = c('Fix_2_V1', "Fix_2_V2", "Fix_2_V3"),
             stringsAsFactors = FALSE),
  'd', 
  temporary = TRUE, overwrite = TRUE)
dplyr::mutate(
  d,
  valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                    fix1, valuesA),
  valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                    fix2, valuesA),
  valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                    fix1, valuesB),
  valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                    fix2, valuesB))
#> # Source: lazy query [?? x 6]
#> # Database: sqlite 3.19.3 [:memory:]
#>   valuesA  canUseFix1 fix1     canUseFix2 fix2     valuesB 
#>   <chr>         <int> <chr>         <int> <chr>    <chr>   
#> 1 A                 1 Fix_1_V1          0 Fix_2_V1 B       
#> 2 Fix_1_V2          1 Fix_1_V2          0 Fix_2_V2 Fix_1_V2
#> 3 <NA>              0 Fix_1_V3          1 Fix_2_V3 Fix_2_V3
```




