---
title: "Dependencies"
author: "Win-Vector LLC"
date: "11/30/2017"
output: github_document
---

Set up our example.

```{r setup}
packageVersion("dplyr")

my_db <- DBI::dbConnect(RSQLite::SQLite(),
                        ":memory:")
d <- dplyr::copy_to(my_db, 
                    data.frame(valuesA = c("A", NA, NA),
                               valuesB = c("B", NA, NA),
                               canUseFix1 = c(TRUE, TRUE, FALSE),
                               fix1 = c('Fix_1_V1', "Fix_1_V2", "Fix_1_V3"),
                               canUseFix2 = c(FALSE, FALSE, TRUE),
                               fix2 = c('Fix_2_V1', "Fix_2_V2", "Fix_2_V3"),
                               stringsAsFactors = FALSE),
                    'd', 
                    temporary = TRUE, overwrite = TRUE)
knitr::kable(dplyr::collect(d))
```

For our example we are using `canUseFix1*` columns to find which positions
of our `values*` columns can be replaced by the corresponding fix values.
This is a common situation in data processing: where you have a column you
wish to populate from a ordered sequence of alternate sources.

We could write this as nested `ifelse()` or coalesce. But suppose 
we had written the code as below.

```{r badresult}
fixed <- dplyr::mutate(d,
                       valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                                         fix1, valuesA),
                       valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                                         fix2, valuesA),
                       valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                                         fix1, valuesB),
                       valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                                         fix2, valuesB))

fixed ->.;
  dplyr::select(., valuesA, valuesB) ->.;
  dplyr::collect(.) ->.;
  knitr::kable(.)
```

Notice this *silently* failed! It gave a wrong answer, with no indicated error.

The third `valuesA` value remains at `NA` even though it should 
have been repaired by the fix 2 rule.  This is not due to order of statements as
the fix rules were deliberately chosen to apply to disjoint rows.

[`seplyr`](https://winvector.github.io/seplyr/) has a fix/work-around: automatically break up the steps into safe blocks ([announcement](http://www.win-vector.com/blog/2017/11/win-vector-llc-announces-new-big-data-in-r-tools/); here we are using the development [`seplyr`](https://winvector.github.io/seplyr/) `0.5.1` version of [`mutate_se()`](https://winvector.github.io/seplyr/reference/mutate_se.html)).

```{r goodresult2}
library("seplyr")
packageVersion("seplyr")

d %.>% 
  mutate_nse(., 
             valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                               fix1, valuesA),
             valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                               fix2, valuesA),
             valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                               fix1, valuesB),
             valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                               fix2, valuesB),
             mutate_nse_printPlan = TRUE) %.>% 
  select_se(., c("valuesA", "valuesB")) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

`seplyr` used safe statement re-ordering to break the calculation into the minimum
number of blocks/groups that have no in-block dependencies between statements (note
this is more efficient that merely introducing a new mutate each first time a new
value is used).

We can slow that down and see how the underlying planning
functions break the assignments down into a small
number of safe blocks (here we are using the development [`wrapr`](https://winvector.github.io/wrapr/) `1.0.2` function
[`qae()`](https://winvector.github.io/wrapr/reference/qae.html)).

```{r goodresult2s}
packageVersion("wrapr")

steps <- qae(valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                               fix1, valuesA),
             valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                               fix2, valuesA),
             valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                               fix1, valuesB),
             valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                               fix2, valuesB))
print(steps)
plan <- partition_mutate_se(steps)
print(plan)

d %.>% 
  mutate_seb(., plan) %.>% 
  select_se(., c("valuesA", "valuesB")) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

For more on [`seplyr`](https://winvector.github.io/seplyr/) [please start here](http://winvector.github.io/FluidData/IntroductionToSeplyr.html).

