---
title: "Dependencies"
author: "Win-Vector LLC"
date: "11/30/2017"
output: github_document
---

This is an example of an erroneous calculation in `dplyr` `0.7.4`, likely
arising from expression to expression dependencies (which [we have written about
before](http://www.win-vector.com/blog/2017/09/my-advice-on-dplyrmutate/)).

First we set up our example.

```{r setup}
packageVersion("dplyr")

my_db <- DBI::dbConnect(RSQLite::SQLite(),
                        ":memory:")
d <- dplyr::copy_to(my_db, 
                    data.frame(valuesA = c("A", NA, NA),
                               valuesB = c("B", NA, NA),
                               canUseFix1 = c(TRUE, TRUE, FALSE),
                               fix1 = c('Fix_1_V1', "Fix_1_V2", "Fix_1_V3"),
                               canUseFix2 = c(FALSE, FALSE, TRUE),
                               fix2 = c('Fix_2_V1', "Fix_2_V2", "Fix_2_V3"),
                               stringsAsFactors = FALSE),
                    'd', 
                    temporary = TRUE, overwrite = TRUE)
knitr::kable(dplyr::collect(d))
```

For our example we are using `canUseFix1*` columns to find which positions
of our `values*` columns can be replaced by the corresponding fix values.
This is a common situation in data processing: where you have a column you
wish to populate from a ordered sequence of alternate sources.

We could write this as nested `ifelse()` or coalesce. But suppose 
we had written the code as below.

```{r badresult}
fixed <- dplyr::mutate(d,
                       valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                                         fix1, valuesA),
                       valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                                         fix2, valuesA),
                       valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                                         fix1, valuesB),
                       valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                                         fix2, valuesB))

fixed ->.;
  dplyr::select(., valuesA, valuesB) ->.;
  dplyr::collect(.) ->.;
  knitr::kable(.)
```

Notice this *silently* failed! It gave a wrong answer, with no indicated error.

The third `valuesA` value remains at `NA` even though it should 
have been repaired by the fix 2 rule.  This is not due to order of statements as
the fix rules were deliberately chosen to apply to disjoint rows.

Looking further we see `dplyr` seem to generate incomplete `SQL` (not all the 
intended transforms seem to survive the translation, notice there are 3 `CASE WHEN`
statements in the generate `SQL, not 4):

```{r badsql}
d  ->.;
  dplyr::mutate(.,
                valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                                  fix1, valuesA),
                valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                                  fix2, valuesA),
                valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                                  fix1, valuesB),
                valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                                  fix2, valuesB)) ->.;
  dplyr::show_query(.)
```


This has been filed as a [`dplyr` issue](https://github.com/tidyverse/dplyr/issues/3223), and 
the issue might be avoided by an [upcoming code change](https://github.com/tidyverse/dbplyr/commit/36a44cd4b5f70bc06fb004e7787157165766d091) (though we have not confirmed the change will fix this particular issue, and question if the change generates efficient `SQL`).

For our recommended current work-around, please see [here](http://winvector.github.io/FluidData/DplyrDependencies.html).

