---
title: "Advanced dplyr Quiz"
author: "John Mount"
date: "6/24/2017"
output:
  md_document:
    variant: markdown_github
---


# Advanced dplyr Quiz

Being able to effectively perform meaningful work *using* [`R`](https://www.r-project.org) programming involves being able to both know how various packages work and anticipate package method outcomes in basic situations.  Any mismatch there (be it a knowledge gap in the programmer, an implementation gap in a package, or a difference between programmer opinion and package doctrine) can lead to confusion, bugs and incorrect results.

Below is our advanced [`dplyr`](https://CRAN.R-project.org/package=dplyr) quiz. Can you anticipate the result of each of the example operations?  Can you anticipate which commands are in error and which are valid `dplyr`?

Or another phrasing: here are our notes on `dplyr` corner-cases.  You may not need to know how any of these work (it is often good to avoid corner-cases), but you should at least be confident you are avoiding the malformed ones.

# Start 

With the current version of `dplyr` in mind, please anticipate the result of each example command. Note: we don't claim all of the examples below are correct `dplyr` code. However, effective programming requires knowledge of what happens in some incorrect cases (at least knowing which throw usable errors, and which perform quiet mal-calculations).

```{r config, echo=FALSE}
DORUN <- FALSE
```

```{r setup, eval=DORUN, echo=DORUN}
# Show versions we are using.
# devtools::install_github('tidyverse/dplyr')
# devtools::install_github('tidyverse/dbplyr')
# devtools::install_github('rstats-db/RSQLite')
# devtools::install_github('tidyverse/rlang')
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
packageVersion("dbplyr")
packageVersion("RSQlite")
packageVersion("rlang")
packageVersion("magrittr")
base::date()
```

Now for the examples/quiz.

Please take a moment and write down your answers before moving on to the [solutions](https://github.com/WinVector/Examples/blob/master/dplyr/dplyrQuiz_solutions.md).  This should give you a much more open mind as to what constitutes "surprising behavior." You can also run the quiz yourself by downloading and knitting the [source document](https://github.com/WinVector/Examples/blob/master/dplyr/dplyrQuiz.Rmd).  

Please keep in mind while "you never want errors" you do sometimes want exceptions (which are unfortunately called "`Error:`" in `R`).   Exceptions are an important way of stopping off-track computation and preventing later incorrect results. Exceptions can often be the desired outcome of a malformed calculation.

# Local data.frames

## Column selection

```{r cselect, eval=DORUN}
data.frame(x = 1) %>% select('x')

y <- 'x' # value used in later examples

data.frame(x = 1) %>% select(y)
data.frame(x = 1, y = 2) %>% select(y)
```

(From [dplyr 2904](https://github.com/tidyverse/dplyr/issues/2904).)


## Piping into different targets (functions, blocks expressions):

`magrittr` pipe details:

```{r pipesdots, error=TRUE, eval=DORUN}
data.frame(x = 1)  %>%  { bind_rows(list(., .)) }
data.frame(x = 1)  %>%    bind_rows(list(., .))
data.frame(x = 1)  %>%  ( bind_rows(list(., .)) )
```

Same with [Bizarro Pipe](https://cran.r-project.org/web/packages/replyr/vignettes/BizarroPipe.html):

```{r pipesdotsb, eval=DORUN}
data.frame(x = 1)  ->.;  { bind_rows(list(., .)) }
data.frame(x = 1)  ->.;    bind_rows(list(., .))
data.frame(x = 1)  ->.;  ( bind_rows(list(., .)) )
```


## enquo rules


```{r equor, error=TRUE, eval=DORUN}
(function(x) select(data.frame(x = 1), !!enquo(x)))(x)
(function(x) data.frame(x = 1) %>% select(!!enquo(x)))(x)
```

(From [dplyr 2726](https://github.com/tidyverse/dplyr/issues/2726).)


# Databases

Setup:

```{r dsetup, eval=DORUN}
# values used in later examples
db <- DBI::dbConnect(RSQLite::SQLite(), 
                     ":memory:")
dL <- data.frame(x = 3.077, 
                k = 'a', 
                stringsAsFactors = FALSE)
dR <- dplyr::copy_to(db, dL, 'dR')
```

## nrow()

```{r nrow, eval=DORUN}
nrow(dR)
```

(From [dplyr 2871](https://github.com/tidyverse/dplyr/issues/2871).)

## union_all()

```{r union_all, error=TRUE, eval=DORUN}
union_all(dR, dR)
union_all(dR, head(dR))
```

(From [dplyr 2858](https://github.com/tidyverse/dplyr/issues/2858).)


## mutate_all funs()

```{r mutate_funs_db, error=TRUE, eval=DORUN}
dR %>% mutate_all(funs(round(., 2)))
dR %>% mutate_all(funs(round(., digits = 2)))
```

(From [dplyr 2890](https://github.com/tidyverse/dplyr/issues/2890) and [dplyr 2908](https://github.com/tidyverse/dplyr/issues/2908).)

## rename

```{r rename_db, error=TRUE, eval=DORUN}
dR %>% rename(x2 = x) %>% rename(k2 = k)
dR %>% rename(x2 = x, k2 = k)
```

(From [dplyr 2860](https://github.com/tidyverse/dplyr/issues/2860).)

# Conclusion

The above quiz is really my working notes on corner-cases to avoid.  Not all of these are worth fixing. In many cases you can and should re-arrange your `dplyr` pipelines to avoid triggering the above cases.  But to do that, you have to know what to avoid (hence the notes).  I may or may not keep these up to date depending on the utility of such a list going forward.


