---
title: "Advanced dplyr Quiz"
author: "John Mount"
date: "6/24/2017"
output:
  md_document:
    variant: markdown_github
---


# Advanced dplyr Quiz

Being able to effectively do work through programming involves being able to both know how various packages work and anticipate package method outcomes in basic situations.  Any mis-match here (be it a knowledge gap in the programmer, or an implementation gap in the package) can lead to bugs and confusion.

Below is our advanced `dplyr` quiz: can you anticipate the result of each of the example operations?

# Start 

With the following version of `dplyr` please figure out the result of each example command. Note: we don't claim all of the below is correct `dplyr` code, however programming requires knowledge of what happens in incorrect cases (for examples which throw usable errors, and which perform subtle mal-calculations).

First let's start up the latest version of all the packages we are using.

```{r setup}
# devtools::install_github('tidyverse/dplyr')
# devtools::install_github('tidyverse/dbplyr')
# devtools::install_github('rstats-db/RSQLite')
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
packageVersion("dbplyr")
packageVersion("RSQlite")
packageVersion("magrittr")
base::date()
DORUN <- FALSE
```

Now the examples/quiz.  Please take a moment and write down your answer before moving on to the solutions.

# Local data.frames

## Column selection

```{r cselect, eval=DORUN}
data.frame(x = 1) %>% select('x')

y <- 'x'

data.frame(x = 1) %>% select(y)
data.frame(x = 1, y = 2) %>% select(y)
```

(From [dplyr 2904](https://github.com/tidyverse/dplyr/issues/2904).)


## Piping into different targets (functions, blocks expressions):

`magrittr` pipe details:

```{r pipesdots, error=TRUE, eval=DORUN}
data.frame(x = 1)  %>%  { bind_rows(list(., .)) }
data.frame(x = 1)  %>%    bind_rows(list(., .))
data.frame(x = 1)  %>%  ( bind_rows(list(., .)) )
```

Same with [Bizarro Pipe](https://cran.r-project.org/web/packages/replyr/vignettes/BizarroPipe.html):

```{r pipesdotsb, eval=DORUN}
data.frame(x = 1)  ->.;  { bind_rows(list(., .)) }
data.frame(x = 1)  ->.;    bind_rows(list(., .))
data.frame(x = 1)  ->.;  ( bind_rows(list(., .)) )
```



## enquo rules


```{r equor, error=TRUE, eval=DORUN}
(function(x) select(data.frame(x = 1), !!enquo(x)))(x)
(function(x) data.frame(x = 1) %>% select(!!enquo(x)))(x)
```

(From [dplyr 2726](https://github.com/tidyverse/dplyr/issues/2726).)


# Databases

Setup:

```{r dsetup, eval=DORUN}
db <- DBI::dbConnect(RSQLite::SQLite(), 
                     ":memory:")
dR <- dplyr::copy_to(db,
                     data.frame(x = 3.077, 
                                k = 'a', 
                                stringsAsFactors = FALSE), 
                     'dR')
```

## nrow()

```{r nrow, eval=DORUN}
nrow(dR)
```

(From [dplyr 2871](https://github.com/tidyverse/dplyr/issues/2871).)

## union_all()

```{r union_all, error=TRUE, eval=DORUN}
union_all(dR, dR)
union_all(dR, head(dR))
```

(From [dplyr 2858](https://github.com/tidyverse/dplyr/issues/2858).)


## mutate_all funs()

```{r mutate_funs_db, error=TRUE, eval=DORUN}
dR %>% mutate_all(funs(round(., 2)))
dR %>% mutate_all(funs(round(., digits = 2)))
```

(From [dplyr 2890](https://github.com/tidyverse/dplyr/issues/2890).)

## rename

```{r rename_db, error=TRUE, eval=DORUN}
dR %>% rename(x2 = x) %>% rename(k2 = k)
dR %>% rename(x2 = x, k2 = k)
```

(From [dplyr 2860](https://github.com/tidyverse/dplyr/issues/2860).)
