---
title: "Better Grouped Summaries in dplyr"
author: "John Mount"
date: "7/12/2017"
output:
  md_document:
    variant: markdown_github
---

For [`R`](https://cran.r-project.org) [`dplyr`](https://CRAN.R-project.org/package=dplyr) users one of the promises of the new [`rlang`/`tidyeval`](https://CRAN.R-project.org/package=rlang) system is an improved ability to [program over `dplyr`](http://dplyr.tidyverse.org/articles/programming.html) itself. In particular to add new verbs that encapsulate previously compound steps into better self-documenting atomic steps.

Let's take a look at this capability.

First let's start `dplyr`.

```{r start}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
```

A `dplyr` pattern that I have seen used often is the "`group_by() %>% mutate()`" pattern.  This historically has been shorthand for a "`group_by() %>% summarize()`" followed by a `join()`.  It is easiest to show by example.

The following code:

```{r ex1l}
mtcars %>% 
  group_by(cyl, gear) %>%
  mutate(group_mean_mpg = mean(mpg), 
            group_mean_disp = mean(disp)) %>% 
  select(cyl, gear, mpg, disp, 
         group_mean_mpg, group_mean_disp) %>%
  head()
```

is taken to be shorthand for:

```{r ex1s}
mtcars %>% 
  group_by(cyl, gear) %>%
  summarize(group_mean_mpg = mean(mpg), 
            group_mean_disp = mean(disp)) %>% 
  left_join(mtcars, ., by = c('cyl', 'gear')) %>%
  select(cyl, gear, mpg, disp, 
         group_mean_mpg, group_mean_disp) %>%
  head()
```

The advantages of the shorthand are:

 * The analyst only has to specify the grouping column once.
 * The data (`mtcars`) enters the pipeline only once.
 * The analyst doesn't have to start thinking about joins immediately.
 
Frankly I've never liked the shorthand. I feel it is a "magic extra" that a new 
user would have no way of anticipating from common use of `group_by()` and `summarize()`.
I very much like the idea of wrapping this important common use case into a single
verb.  Adjoining "windowed" or group-calculated columns is a common and important
step in analysis, and well worth having its own verb.

Below is our attempt at elevating this pattern into a packaged verb.


```{r fn}
#' Simulate the group_by/mutate pattern with an explicit summarize and join.
#' 
#' Group a data frame by the groupingVars and compute user summaries on 
#' this data frame (user summaries specified in ...), then join these new
#' columns back into the original data and return to the user.
#' This works around https://github.com/tidyverse/dplyr/issues/2960 .
#' And it is a demonstration of a higher-order dplyr verb.
#' Author: John Mount, Win-Vector LLC.
#' 
#' @param d data.frame
#' @param groupingVars character vector of column names to group by.
#' @param ... list of dplyr::mutate() expressions.
#' @value d with grouped summaries added as extra columns
#' 
#' @examples
#' 
#' add_group_summaries(mtcars, 
#'                     c("cyl", "gear"), 
#'                     group_mean_mpg = mean(mpg), 
#'                     group_mean_disp = mean(disp)) %>%
#'   head()
#' 
#' @export
#' 
add_group_summaries <- function(d, groupingVars, ...) {
  # convert char vector into spliceable vector
  groupingSyms <- rlang::syms(groupingVars)
  dg <- ungroup(d) # just in case
  dg <- group_by(dg, !!!groupingSyms)
  ds <- summarize(dg, ...)
  # work around https://github.com/tidyverse/dplyr/issues/2963
  ds <- ungroup(ds)
  left_join(d, ds, by= groupingVars)
}
```

This works as follows:

```{r exlocal}
mtcars %>% 
  add_group_summaries(c("cyl", "gear"), 
                      group_mean_mpg = mean(mpg), 
                      group_mean_disp = mean(disp)) %>%
  select(cyl, gear, mpg, disp, 
         group_mean_mpg, group_mean_disp) %>%
  head()
```

And this also works on database-backed `dplyr` data (which the shorthand currently does not, please see [`dplyr` 2887 issue](https://github.com/tidyverse/dplyr/issues/2887) and [`dplyr` issue 2960](https://github.com/tidyverse/dplyr/issues/2960)).

```{r exdb1, error=TRUE}
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, mtcars)
mtcars2 <- tbl(con, "mtcars")

mtcars2 %>% 
  group_by(cyl, gear) %>%
  mutate(group_mean_mpg = mean(mpg), 
         group_mean_disp = mean(disp))
```

```{r exdb2}
mtcars2 %>% 
  add_group_summaries(c("cyl", "gear"), 
                      group_mean_mpg = mean(mpg), 
                      group_mean_disp = mean(disp)) %>%
  select(cyl, gear, mpg, disp, 
         group_mean_mpg, group_mean_disp) %>%
  head()
```

Another great verb in this style is `group_summarize`:

```{r fn2}
#' group_by and summarize as an atomic action.
#' 
#' Group a data frame by the groupingVars and compute user summaries on 
#' this data frame (user summaries specified in ...).  Enforces the 
#' good dplyr pipeline design principle of keeping group_by and
#' summarize close together.
#' Author: John Mount, Win-Vector LLC.
#' 
#' @param d data.frame
#' @param groupingVars character vector of column names to group by.
#' @param ... list of dplyr::mutate() expressions.
#' @value d summarized by groups
#' 
#' @examples
#' 
#' group_summarize(mtcars, 
#'                     c("cyl", "gear"), 
#'                     group_mean_mpg = mean(mpg), 
#'                     group_mean_disp = mean(disp)) %>%
#'   head()
#' 
#' @export
#' 
group_summarize <- function(d, groupingVars, ...) {
  # convert char vector into spliceable vector
  groupingSyms <- rlang::syms(groupingVars)
  dg <- ungroup(d) # just in case
  dg <- group_by(dg, !!!groupingSyms)
  ds <- summarize(dg, ...)
  # work around https://github.com/tidyverse/dplyr/issues/2963
  ungroup(ds)
}
```

```{r exg}
mtcars2 %>% 
  group_summarize(c("cyl", "gear"), 
                  group_mean_mpg = mean(mpg), 
                  group_mean_disp = mean(disp)) %>%
  select(cyl, gear,
         group_mean_mpg, group_mean_disp)
```


