---
title: "Please Consider Using `wrapr::let()` for Replacement Tasks"
author: "John Mount"
date: "6/26/2017"
output:
  md_document:
    variant: markdown_github
---

(author: John Mount)

# Please Consider Using `wrapr::let()` for Replacement Tasks

From [`dplyr` issue 2916](https://github.com/tidyverse/dplyr/issues/2916).

The following *appears* to work.

```{r ex1}
suppressPackageStartupMessages(library("dplyr"))

COL <- "homeworld"
starwars %>%
  group_by(.data[[COL]]) %>%
  head(n=1)
```

Though notice it reports the grouping is by "`COL`", not by "`homeworld`".  Also the data set now has `14` columns, not the original `13` from the `starwars` data set.

And this seemingly similar variation (currently) throws an exception:

```{r ex1b, error=TRUE}
homeworld <- "homeworld"

starwars %>%
  group_by(.data[[homeworld]]) %>% 
  head(n=1) 
```

I know this will cost me what little community good-will I might have left (after already having raised this, unsolicited, many times), but *please* consider using our package `wrapr::let()` for tasks such as the above.

```{r ex2}
library("wrapr")

let(
  c(COL = "homeworld"),
  
  starwars %>%
    group_by(COL) %>%
    head(n=1)
)

let(
  c(homeworld = "homeworld"),
  
  starwars %>%
    group_by(homeworld) %>% 
    head(n=1)
)
```

Some explanation can be found [here](http://www.win-vector.com/blog/2017/06/wrapr-implementation-update/).
