---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r localExample}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")
packageVersion("sparklyr")

# function to order columns and rows for unambiguous presentation of results
present <- . %>%
  ungroup() %>%
  select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, Species) %>%
  arrange(Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)

# local data example of 
# dplyr::group_by() %>% dplyr::do()
f <- . %>% 
  arrange(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) %>%
  head(2)

iris %>% 
  group_by(Species) %>% 
  do(f(.)) %>% 
  present()
```

Notice on spark the same code return empty data.

```{r Spark, error=TRUE}
# try it again on Spark

sc <- sparklyr::spark_connect(version='2.0.2', 
                              master = "local")
diris <- copy_to(sc, iris, 'diris')
head(diris)

# function to order columns and rows for unambiguous presentation of results
presentg <- . %>%
  ungroup() %>%
  select(Sepal_Length, Sepal_Width, Petal_Length, Petal_Width, Species) %>%
  arrange(Species, Sepal_Length, Sepal_Width, Petal_Length, Petal_Width)

# function with column names matching Spark column names
f2 <- . %>% 
  arrange(Sepal_Length, Sepal_Width, Petal_Length, Petal_Width) %>%
  head(2)

diris %>% 
  group_by(Species) %>% 
  do(f2(.))
```

If the function we are applying is compatible with `dplyr::group_by()`, then we can try a direct application of it per group in as follows.  In this case it does not work, as `head()` is not "group aware":

```{r groubyf, error=TRUE}
diris %>%
  group_by(Species) %>%
  f2 %>%
  presentg()
```


`dplyr::slice()` also fails.

```{r SparkSlice, error=TRUE}
f2s <- . %>% 
  arrange(Sepal_Length, Sepal_Width, Petal_Length, Petal_Width) %>%
  slice(1:2)

diris %>% 
  group_by(Species) %>% 
  do(f2s(.))

diris %>%
  group_by(Species) %>%
  arrange(Sepal_Length, Sepal_Width, Petal_Length, Petal_Width) %>%
  slice(1:2)
```

For a moderate number of groups `replyr::replyr_split()` can do the job.

```{r replyr}
# try it with replyr
# devtools::install_github('WinVector/replyr')
library("replyr")
packageVersion("replyr")

# gapply extract method, only appropriate for small number of
# groups, could use 'group_by', but that requires f2
# respect groups (head() does not and slice() isn't available
# on this verion of Spark/Sparklyr)
diris %>% 
  gapply('Species', partitionMethod='extract', f2) %>%
  presentg()

# Or in separate stages
diris %>% 
  replyr_split('Species') %>%
  lapply(f2) %>%
  replyr_bind_rows() %>%
  presentg()
```

For a large number of groups you must find a way to write your transform entirely in "group_by" compatible verbs.  For this application it appears `row_number()` and `filter()` can be used to re-write our function.

```{r rownumber}
# group compatible version of function.
f3g <- . %>%
  arrange(Sepal_Length, Sepal_Width, Petal_Length, Petal_Width) %>%
  filter(between(row_number(), 1, 2))

diris %>% 
  group_by(Species) %>% 
  f3g() %>%
  presentg()
```

Again, you no longer can use `do()` notation:

```{r rownumberdo, error=TRUE}
diris %>% 
  group_by(Species) %>% 
  do(f3g(.))
```

```{r cleanup}
sparklyr::spark_disconnect(sc)
rm(list=ls())
gc()
```
