---
title: "simple_betting"
output: github_document
---

I would like to write about [Thorp](https://en.wikipedia.org/wiki/Edward_O._Thorp) style card counting using the [Kelly betting criterion](https://en.wikipedia.org/wiki/Kelly_criterion).

The example I will explore is: betting if the next card in a deck is red or black. For a single deck version of this game,
one can vary ones bets to exploit knowledge of which cards remain in the deck.

A video of me playing the game can be found [here](https://youtu.be/6xhjbgREGDA).


<iframe width="560" height="315" src="https://www.youtube.com/embed/6xhjbgREGDA" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Let's show how we develop such a betting stategy.

First we import packages and initialize state.

```{r}
library(cdata)
library(ggplot2)
library(rqdatatable)

set.seed(2021)
```

We define our betting function. This function takes `na`, the number of `A`-cards remaining, `nb` the number of `B`-cards
remaining, and `stake` our remaining dollars. It returns the Kelly bet, which in this case is `2 (pa - 1/2)` where `pa` is the
probability of drawing an `A`-card, which in turn equals `na / (na + nb)`.

```{r}
bet_Kelly <- function(na, nb, stake) {
  sign <- 1
  if(na < nb) {
    sign <- -1
    tmp <- na
    na <- nb
    nb <- tmp
  }
  ideal_bet_fraction <- 0
  bet <- 0
  if((stake > 1) && (na != nb) && ((na + nb) > 0)) {
     ideal_bet_fraction <- 2 * (na / (na + nb) - 1/2)
     bet <- min(stake - 1, round(stake * ideal_bet_fraction))
  }
  return(list(bet = sign * bet, ideal_bet_fraction = sign * ideal_bet_fraction))
}
```

With this function we can
write down our betting card:
the percent of stake bets as a function of remaining `a`s and `b`s.

```{r}
n_seq <- seq(0, 26, by = 2)
strategy_table <- matrix(data = NA, nrow = length(n_seq), ncol = length(n_seq))
for(nai in seq_len(length(n_seq))) {
  na <- n_seq[nai]
  for(nbi in seq_len(length(n_seq))) {
    nb <- n_seq[nbi]
    strategy_table[nai, nbi] = bet_Kelly(na, nb, 100)$bet
  }
}
rownames(strategy_table) <- paste0('na~', n_seq)
colnames(strategy_table) <- paste0('nb~', n_seq)
knitr::kable(strategy_table)
```

Let's see this strategy in (simulated) action.

We define evaluation functions that draw the cards and simulate the play.

```{r}
simulate_draws <-  function(cards, steps = cards, replace = FALSE) {
  d <- data.frame(
    step = seq_len(steps + 1) - 1,
    na = floor(cards/2)
  )
  d$nb <- cards - d$na
  d$draw_is_a <- NA
  for(i in seq_len(steps)) {
    if(!replace) {
      d$draw_is_a[i] <- runif(n = 1) <= d$na[i] / (d$na[i] + d$nb[i])
    } else {
      d$draw_is_a[i] <- runif(n = 1) <= 0.5
    }
    if(d$draw_is_a[i]) {
      d$na[i+1] <- d$na[i] - 1
      d$nb[i+1] <- d$nb[i]
    } else {
      d$na[i+1] <- d$na[i]
      d$nb[i+1] <- d$nb[i] - 1
    }
  }
  return(d)
}

simulate_play <- function(
  d,
  stake,
  bet_fn) {
  d$stake <- stake
  d$bet <- NA
  d$ideal_bet_fraction <- NA
  for(i in seq_len(nrow(d)-1)) {
    beti <- bet_fn(na = d$na[i], nb = d$nb[i], stake = d$stake[i])
    d$bet[i] <- beti$bet
    d$ideal_bet_fraction[i] <- beti$ideal_bet_fraction
    if(d$bet[i] != 0) {
      if(abs(d$bet[i]) > d$stake[i]) {
        stop("bet more than stake")
      }
      d$stake[i+1] <- d$stake[i] + 
        sign(d$bet[i]) * ifelse(d$draw_is_a[i], 1, -1) * abs(d$bet[i])
    } else {
      d$stake[i+1] <- d$stake[i]
    }
  }
  return(d)
}
```

We show a couple of example plays.

```{r}
simulate_draws(cards = 52) %.>%
  simulate_play(., stake = 100, bet_fn = bet_Kelly) %.>%
  knitr::kable(.)
```

```{r}
simulate_draws(cards = 52) %.>%
  simulate_play(., stake = 100, bet_fn = bet_Kelly) %.>%
  knitr::kable(.)
```


Our Thorp-style count based approximate betting rule can be found by estimating
what fraction of a stake simulates the exact Kelly rule on average. I chose
to use a linear model via `lm()` to get the approximation.


```{r}
# collect runs
simulations <- lapply(
  seq_len(1000),
  function(i) {
    v <- simulate_draws(cards = 52) %.>%
    simulate_play(., stake = 100, bet_fn = bet_Kelly)
    return(v)
  })

# buld observed delta count to bet tables
obs <- lapply(
  simulations,
  function(s) {
    return(data.frame(na = s$na, nb = s$nb, delta = s$na - s$nb, bet = s$bet))
  })
obs <- do.call(rbind, obs)
model <- lm(bet ~ 0 + delta, data = obs)
mult <- round(model$coefficients[[1]])
mult
```

The result is: bet about `r mult` percent of your stake for each count of imbalance (explained in the
[video](https://youtu.be/6xhjbgREGDA)). In the video we use a lower 5 percent target stake bet. Under-betting
is less dangerous than over-betting, and the under-betting strategy "half Kelly" is actually quite common in 
finance.

We instantiate the simplified betting rule, and some over and under betting variations.

```{r}
bet_simple <- function(stake_mult) {
  force(stake_mult)
  return(
    function(na, nb, stake) {
      sign <- 1
      if(na < nb) {
        sign <- -1
        tmp <- na
        na <- nb
        nb <- tmp
      }
      ideal_bet_fraction <- 0
      bet <- 0
      if((stake > 1) && (na != nb) && ((na + nb) > 0)) {
         ideal_bet_fraction <- stake_mult * (na - nb) / 100
         bet <- min(stake - 1, round(stake * ideal_bet_fraction))
      }
      return(list(bet = sign * bet, ideal_bet_fraction = sign * ideal_bet_fraction))
    })
}

bet_simple_est_mult <- bet_simple(mult)
bet_simple_10 <- bet_simple(10)
bet_bet_simple_10 <- bet_simple(5)
```

We show a couple of example plays.

```{r}
simulate_draws(cards = 52) %.>%
  simulate_play(., stake = 100, bet_fn = bet_simple_est_mult) %.>%
  knitr::kable(.)
```

```{r}
simulate_draws(cards = 52) %.>%
  simulate_play(., stake = 100, bet_fn = bet_simple_est_mult) %.>%
  knitr::kable(.)
```

And we define some over and under Kelly betting functions.

```{r}
lambda_Kelly <- function(lambda) {
  force(lambda)
  return(
    function(na, nb, stake) {
      bk <- bet_Kelly(na = na, nb = nb, stake = stake)
      return(list(
        bet = sign(bk$bet) * min(stake - 1, round(lambda * abs(bk$bet))),
        ideal_bet_fraction = bk$ideal_bet_fraction))
    })
}
```

We now simulate all the betting strategies on the same card draws.

```{r}
fns = list(
  # bet_Kelly = bet_Kelly,
  bet_simple_est_mult = bet_simple_est_mult,
  bet_simple_10 = bet_simple(10),
  bet_simple_5 = bet_simple(5),
  f_0.50_Kelly = lambda_Kelly(0.5),
  f_0.75_Kelly = lambda_Kelly(0.75),
  f_1.00_Kelly = lambda_Kelly(1.0),
  f_1.50_Kelly = lambda_Kelly(1.5),
  f_2.00_Kelly = lambda_Kelly(2.0)
)

runs <- lapply(
  seq_len(1000),
  function(i) {
    d <- simulate_draws(cards = 52)
    sims <- lapply(
      fns,
      function(f) {
        simulate_play(d, stake = 100, bet_fn = f)
      })
    return(as.data.frame(lapply(
      sims,
      function(d) {
        d$stake[nrow(d)]
      })))
  })
runs <- do.call(rbind, runs)

summary(runs)
```

WE plot the distribution of results.

```{r}
cols <- colnames(runs)

runs %.>%
  project_se(
    .,
    paste0('Sharpe_ratio_', cols, ' := mean(', cols, ') / sd(', cols, ')'),
    grouping = c()) %.>%
  knitr::kable(.)

# prepare data to plot
# some of the ideas are here:
# https://win-vector.com/2021/02/20/plotting-multiple-curves-in-python/
plot_frame <- pivot_to_blocks(
  runs, 
  nameForNewKeyColumn = 'strategy', 
  nameForNewValueColumn = 'value', 
  columnsToTakeFrom = cols)

ggplot(
  data = plot_frame,
  aes(x = value, color = strategy)) +
  geom_density() + 
  geom_vline(xintercept = 100, linetype = 2) + 
  xlim(min(plot_frame$value), quantile(plot_frame$value, 0.99)) +
  ggtitle("return distribution by betting strategy")
```

Look at expected log-returns, what Kelly is designed to optimize.

```{r}
runs_ln <- runs %.>% 
  extend_se(., 
            paste0('ln_', cols, ' := log(', cols, ')')) %.>%
  drop_columns(., cols)

summary(runs_ln)
```


Simulate runs where sampling is with replacement, so no
profitable strategy is possible.

```{r}
bad_runs <- lapply(
  seq_len(1000),
  function(i) {
    d <- simulate_draws(cards = 52, replace = TRUE)  # break relation of count to probability
    sims <- lapply(
      fns,
      function(f) {
        simulate_play(d, stake = 100, bet_fn = f)
      })
    return(as.data.frame(lapply(
      sims,
      function(d) {
        d$stake[nrow(d)]
      })))
  })
bad_runs <- do.call(rbind, bad_runs)

summary(bad_runs)
```

Notice we confirm the strategies do not work in this case.

