---
title: "Prediction Plot"
output: github_document
---

See if we can infer distributional facts from the empirical ROC plot.

It looks like the prevalence is not obvious from the empirical plot, though it is where the parameterization achieves slope 1, which is also where the distance off the diagonal is maximized.

```{r, message=FALSE, warning=FALSE}
library(wrapr)
library(ggplot2)
library(WVPlots)
```

```{r}
unpack[
  train_p,
  test_p
] <- readRDS('predictions.RDS') # from https://github.com/WinVector/vtreat/tree/main/Examples/KDD2009
```

```{r}
ROCPlot(
  test_p,
  xvar = 'xgboost',
  truthVar = 'churn',
  truthTarget = TRUE,
  title = 'model on test',
  add_convex_hull = TRUE)
```

Use the method of "A Single Parameter Family Characterizing Probability Model Performance" to identify the ROC curve from the restricted family.

First get `a, b` shape family. We are considering curves of the form `(a = c * a1, b = c * b1)` for positive `c`.

```{r}
(prevalence <- mean(test_p$churn))
```

```{r}
(a1 <- 1)
(b1 <- a1 * (1 - prevalence) / prevalence)
a1 / (a1 + b1)
```

Now get the ideal AUC seen.

```{r}
calc_area <- function(x, y) {
  n <- nrow(data)
  dx <- abs(x[-length(x)] - x[-1])
  my <- (y[-length(x)] + y[-1])/2
  sum(dx*my)
}
```

```{r}
roc_curve <- sigr::build_ROC_curve(
  modelPredictions = test_p$xgboost, 
  yValues = test_p$churn)

idxs <- sort(unique(chull(
  x = c(1 - roc_curve$Specificity, 1),
  y = c(roc_curve$Sensitivity, 0))))
idxs <- idxs[-length(idxs)]
convex_hull <- roc_curve[idxs, ]
target_area <- calc_area(
  x = 1 - convex_hull$Specificity, 
  y = convex_hull$Sensitivity)

target_area
```


```{r}
mk_frame <- function(a, b, eval_points = seq(0, 1, 1e-5)) {
  data.frame(
    what = paste0("a=", format(a, digits = 3), ', b=', format(b, digits = 3)),
    a = a,
    b = b,
    model_score = eval_points,
    false_positive_rate = 1 - pbeta(eval_points, shape1 = a, shape2 = b + 1),
    true_positive_rate = 1 - pbeta(eval_points, shape1 = a + 1, shape2 = b),
    stringsAsFactors = FALSE)
}



mk_plot <- function(data) {
  ggplot(
    data = data,
    mapping = aes(
      x = false_positive_rate, 
      y = true_positive_rate, 
      color = what)) +
    geom_line() + 
    geom_abline(intercept = 0, slope = 1) +
    coord_fixed() +
    scale_color_brewer(palette = "Dark2") +
    ggtitle(paste0(
      "theoretical ROC curve(s)"))
}
```

```{r}
pf <- mk_frame(a1, b1)
mk_plot(pf)
calc_area(
  x = pf$false_positive_rate,
  y = pf$true_positive_rate)
```

```{r}
f <- function(x) {
  a <- x[[1]]
  b <- a * (1 - prevalence) / prevalence
  of <- mk_frame(a = a, b = b)
  fn <- suppressWarnings(approxfun(
    x = of$false_positive_rate, 
    y = of$true_positive_rate))
  evals <- fn(1 - roc_curve$Specificity)
  sum((roc_curve$Sensitivity - evals)^2)
}
(soln <- optimize(f, c(0, 100)))
```

```{r}
(a <- soln$minimum)
(b <- a * (1 - prevalence) / prevalence)

print(a/(a + b))

pf <- mk_frame(a, b)
idx <- which.min(abs(prevalence - pf$model_score))
print(pf[idx, ])

mk_plot(pf) + 
  geom_point(
    data = pf[idx, ],
     mapping = aes(
      x = false_positive_rate, 
      y = true_positive_rate)) +
  geom_abline(
    intercept = pf$true_positive_rate[[idx]] - pf$false_positive_rate[[idx]],
    slope = 1)
              

calc_area(
  x = pf$false_positive_rate,
  y = pf$true_positive_rate)
```

```{r}
ROCPlot(
  test_p,
  xvar = 'xgboost',
  truthVar = 'churn',
  truthTarget = TRUE,
  title = 'model on test, compared to theoretical') + 
  geom_line(
    data = pf,
    mapping = aes(
      x = false_positive_rate, 
      y = true_positive_rate),
    linetype = 3,
    color = "DarkGreen")
```

Fit both parameters, by shape.

```{r}
f2 <- function(x) {
  a <- x[[1]]
  b <- x[[2]]
  of <- mk_frame(a = a, b = b)
  fn <- suppressWarnings(approxfun(
    x = of$false_positive_rate, 
    y = of$true_positive_rate))
  evals <- fn(1 - roc_curve$Specificity)
  sum((roc_curve$Sensitivity - evals)^2)
}

print(f2(c(a, b)))

(soln2 <- optim(
  par = c(a, b),
  fn = f2,
  method = 'L-BFGS-B',
  lower = c(0, 0)))

(a2 <- soln2$par[[1]])
(b2 <- soln2$par[[2]])

print(a2/(a2 + b2))

pf2 <- mk_frame(a2, b2)
idx2 <- which.min(abs(a2/(a2 + b2) - pf2$model_score))
print(pf2[idx2, ])

mk_plot(pf2) + 
  geom_point(
    data = pf2[idx2, ],
     mapping = aes(
      x = false_positive_rate, 
      y = true_positive_rate)) +
  geom_abline(
    intercept = pf2$true_positive_rate[[idx]] - pf2$false_positive_rate[[idx]],
    slope = 1)
              

calc_area(
  x = pf2$false_positive_rate,
  y = pf2$true_positive_rate)

ROCPlot(
  test_p,
  xvar = 'xgboost',
  truthVar = 'churn',
  truthTarget = TRUE,
  title = 'model on test, compared to theoretical2') + 
  geom_line(
    data = pf2,
    mapping = aes(
      x = false_positive_rate, 
      y = true_positive_rate),
    linetype = 3,
    color = "DarkGreen")
```
